---
title: "Analysis of responses from a statewide survey with registered boat owners in Alaska, fall 2018"
author: "Tobias Schwoerer"
date: "November 21, 2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Data import
```{r}
library(jsonlite)
library(geojson)
library(leaflet)
library(dplyr)
library(tidyr)
library(Hmisc)
library(ggplot2)
library(RColorBrewer)
library(stringr)
library(summarytools)
library(stats)
library(leaflet.minicharts)

features <- read.csv("D:/Dropbox/DATA/2018_boat_survey/features.csv", stringsAsFactors = FALSE)
#features <- read.csv("C:/Users/Toby/Dropbox/DATA/2018_boat_survey/features.csv", stringsAsFactors = FALSE)
names(features)[names(features)=="LenghtKM"] <- "km"#renaming misspelled column
freqCoding <- read.csv("reference_tables/freqCoding.csv", stringsAsFactors = FALSE) #this file was created to quickly recode trip frequencies

data <- read.csv("D:/Dropbox/DATA/2018_boat_survey/data.csv", stringsAsFactors = FALSE)
#data <- read.csv("C:/Users/Toby/Dropbox/DATA/2018_boat_survey/data.csv", stringsAsFactors = FALSE)
```

#Clean and mutate features.csv
Each row in this file is associated with a track a respondent entered in the mapping tool of the survey (custom programmed in Qualtrics). First, we recode trip frequencies as stated by respondents in form of an interval. We take the mean of the interval for further analysis. Note, respondents stated trip frequencies for the 2018 summer season (current summer) and answered a hypothetical question how the number of trips would have changed would elodea clog some of the waterbodies they boated on.   
```{r}
#calculating pre elodea trip frequencies
features2 <- features%>%
  left_join(freqCoding,  by = c("trackFrequ"="Coding"))   #recoding trip frequencies
features2$freqPre <- round(rowMeans(features2[c("Lower","Upper")], na.rm=TRUE),digits=0)
names(features2)[names(features2)=="Lower"] <- "LowerPre" #Renaming columns for the lower bound of the frequency interval pre invasion
names(features2)[names(features2)=="Upper"] <- "UpperPre" #Renaming columns for the upper bound of the frequency interval pre invasion

#Calculating post elodea trip frequencies
features3 <- features2 %>%
  left_join(freqCoding,  by = c("trackFre_1"="Coding"))
features3$freqPost <- round(rowMeans(features3[c("Lower","Upper")], na.rm=TRUE), digits=0)
names(features3)[names(features3)=="Lower"] <- "LowerPost" #Renaming columns for the lower bound of the frequency interval post invasion
names(features3)[names(features3)=="Upper"] <- "UpperPost" #Renaming columns for the upper bound of the frequency interval post invasion
```

Calculating annual operating cost per route. Here, we calculate the annual km as the product of the stated trip frequency, round-trip length in km (calculated by ArcGIS), the stated hourly boat operating cost, and the inverse of the assumed average travel speed of 10 miles/h (16km/h). Both pre- and post-invasion costs are calculated. 
```{r}
#hCost: hourly operating cost
features3$hCost <- rowMeans(features3[c("hCostHigh","hCostLow")], na.rm=TRUE) #calculating the hCost interval mean for furthe ranalysis, ignore missing 
features3$hCost[features3$hCost==20000] <- NA  #outliers and missing value
#medianCost <- median(features3$hCost, na.rm=TRUE)  #calculating the median cost per h
#features3$hCost <- with(features3, ifelse(is.na(hCost), medianCost, hCost))  # inserting median cost for missing values
features3$hCost[is.nan(features3$hCost)] <- NA  #R tried to get the mean of the Lower and Upper values but if respondent left blank those are NA, so the mean is also NA

#outliers for hCost, turning into NAs anything above $100/h, one was $500/h which was marine based solely, two others freshwater based solely who were $125 and $150/h (likely entering errors)
features3$hCost <- with(features3, ifelse(hCost>100,NA,hCost))

#km: outliers, setting frequency of trips pre if Up to 4 setting it to 1 if km >300
features3$freqPre <- with(features3, ifelse(km>300 & trackFrequ=="Up to 4",1,freqPre))
#Outliers in trip frequencies
features3$freqPre <- with(features3, ifelse(freqPre>100,NA,freqPre))
features3$freqPost <- with(features3, ifelse(freqPost>100,NA,freqPre))

#rank of route
features3$rank <- with(features3, ifelse(rank==0, 1, rank)) #dealing with ranks that are 0

#YrCostPre and Post: calculating the total km per route then annual cost per route
features3$totalKmPre <- with(features3, km * freqPre)
features3$totalKmPost <- with(features3, km * freqPost)
features3$YrCostPre <- with(features3, totalKmPre * hCost)
features3$YrCostPost <- with(features3, totalKmPost * hCost)
#eliminating one marine route that is an outlier
features3<-features3[!features3$totalKmPre>10000,]

#turning hCost into NAs for YrCostPre exceeding income and few government related routes
features3$hCost <- with(features3, ifelse(YrCostPre>50000,NA,hCost))
#recalculating the above YrCostPre and YrCostPost
features3$totalKmPre <- with(features3, km * freqPre)
features3$totalKmPost <- with(features3, km * freqPost)
features3$YrCostPre <- with(features3, totalKmPre * hCost)
features3$YrCostPost <- with(features3, totalKmPost * hCost)

#adding up number of trips annually for each respondent across all of his/her routes
totalTripsByRes <- features3%>%
  group_by(responseID)%>%
  summarise(totalTripsPre = sum(freqPre),
            totalTripsPost = sum(freqPost),
            routeCount = n())%>%
  select(responseID, totalTripsPre, totalTripsPost, routeCount)
features3 <- features3 %>%
  left_join(totalTripsByRes, by="responseID")
```

Missing data, outliers, and recoding on other variables. Associating each route with either private, commercial, or government related operators. 
```{r}
#INCOME recoding, cleaning, and dealing with missing data on income
incCoding <- read.csv("reference_tables/incCoding.csv", stringsAsFactors = FALSE)
features4 <- features3%>%
  dplyr::left_join(incCoding,  by = "persIncomeBtax")
features4$Income <- as.numeric(as.character(features4$Income))
#imputing missing values for income
#medianIncome <- median(features4$Income, na.rm=TRUE)  #calc. median income
#features4$Income <- with(features4, ifelse(is.na(Income), medianIncome, Income))  #using median income for missing income data

#AGE: questionable data, deleting entries coded "Not applicable"
features4$age <- with(features4, ifelse(age==9 | age=="Not applicable",NA,age))
features4$age <- as.numeric(as.character(features4$age))

#ageBin
features4$ageBin <- with(features4, ifelse(age>20&age<=30,"21-30",ifelse(age>30&age<= 40,"31-40",ifelse(age>40&age<=50,"41-50", ifelse(age>50&age<=60,"51-60", ifelse(age>60&age<=70,"61-70", ifelse(age>70&age<=80, "71-80", ">80")))))))
features4$ageBin <- factor(features4$ageBin, levels=c("21-30", "31-40", "41-50", "51-60", "61-70", "71-80",">80"), labels=c("21-30", "31-40", "41-50", "51-60", "61-70", "71-80",">80")) #turning ageBin into factor

#Clean Drain Dry, coding respondents who did not respond to 
names(features4)[names(features4)=="CleanDrainDry"] <- "Clean"
features4$Clean <- with(features4, ifelse(Clean=="", "Did not report", ifelse(Clean=="half the time","50% of the time",ifelse(Clean=="Every time", "100% of the time",ifelse(Clean=="Never", "0% of the time", Clean)))))
features4$Clean <- factor(features4$Clean, levels=c("Did not report","0% of the time","25% of the time","50% of the time","75% of the time","100% of the time"), labels=c("Did not report","0% of the time","25% of the time","50% of the time","75% of the time","100% of the time")) 

#wasInElodea, clean, recode
features4$wasInElodea<-str_replace_all(features$wasInElodea,"\\s","_")
features4$wasInElodea <- with(features4, ifelse(wasInElodea=="",NA,wasInElodea))
features4$wasInElodea <- with(features4, ifelse(wasInElodea=="I_cannot_remember","Cannot remember",wasInElodea))
features4$wasInElodea <- factor(features4$wasInElodea, levels=c("Yes", "No", "Did not report", "Cannot remember"), labels=c("Yes", "No", "Did not report", "Cannot remember"))

#freshOutside: recoding, turning into factor for graphing
features4$freshOutside <- with(features4, ifelse(freshOutside=="Yes","Yes",ifelse(freshOutside=="No","No","Did not report")))
features4$freshOutside <- factor(features4$freshOutside, levels=c("Yes", "No", "Did not report"), labels=c("Yes", "No", "Did not report"))

#PAX cleaning
features4$pax <- with(features4, ifelse(pax=="none"|pax=="43134"|pax=="",NA,pax))
features4$pax <- as.numeric(as.character(features4$pax))

#USER GROUP, TYPE of OPERATOR 
features4$percentCom <- gsub("[[:punct:]]", " ", features4$percentCom)  #getting rid of % sign in character string
features4$percentCom <- as.numeric(as.character(features4$percentCom))  #converting character to numeric
features4$percentCom <- with(features4, ifelse(is.na(percentCom),0,percentCom/100))
features4$percentGov <- gsub("[[:punct:]]", " ", features4$percentGov)
features4$percentGov <- as.numeric(as.character(features4$percentGov))
features4$percentGov <- with(features4, ifelse(is.na(percentGov),0,percentGov/100))
features4$percentPer <- with(features4, 1-percentGov-percentCom)

#correcting wrong respondent entry on percentGov
features4$percentGov <- with(features4, ifelse(percentPer<0,percentGov-percentCom,percentGov))
features4$percentPer <- with(features4, 1-percentGov-percentCom)

features4$type <- with(features4, ifelse(percentPer>0.5, "personal", ifelse(percentCom>0.5, "commercial", "government")))
#Creating subsets by type
featuresCom <- subset(features4, type=="commercial")
featuresPers <- subset(features4, type=="personal")
featuresGov <- subset(features4, type=="government")
```

Preparing features file for leaflet shiny app
```{r}
#Consolidating levels for several categorical variables which will be used to subset the tracks that are shown. note, leaves the class of these variables as character 
featuresApp <- features4
featuresApp$wasInElodea <- with(featuresApp, ifelse(wasInElodea=="Yes","Yes","No"))
featuresApp$freshOutside <- with(featuresApp, ifelse(freshOutside=="Yes","Yes","No or did not report"))
featuresApp$Clean <- with(featuresApp, ifelse(Clean=="100% of the time","100% of the time","Less than 100% of the time"))

write.csv(featuresApp, "appdir/data/featuresApp.csv" )

#creating geojson file combining data.csv with features4.csv 
#system("C:/Users/Toby/Documents/ANALYSIS_R/boatmovement/appdir/data> python.exe .\map-data-joiner.py .\map.geojson .\features4.csv .\joined-output.geojson")
```

#Clean and mutate data.csv file 
```{r}
#trimming % sign in character string, then converting character to numeric, doing for percentCom and percentGov
data2 <- data
data2$percentCom <- gsub("[[:punct:]]", " ", data2$percentCom)
data2$percentCom <- as.numeric(as.character(data2$percentCom))
data2$percentCom <- with(data2, ifelse(is.na(percentCom),0,percentCom/100))
data2$percentGov <- gsub("[[:punct:]]", " ", data2$percentGov)
data2$percentGov <- as.numeric(as.character(data2$percentGov))
data2$percentGov <- with(data2, ifelse(is.na(percentGov),0,percentGov/100))
data2$percentPer <- with(data2, 1-percentGov-percentCom)

#correcting wrong respondent entry on percentGov
data2$percentGov <- with(data2,ifelse(percentPer<0,percentGov-percentCom,percentGov))
data2$percentPer <- with(data2, 1-percentGov-percentCom)

#renaming columns
names(data2)[names(data2)=="CleanDrainDry"] <- "Clean"

#INCOME
incCoding <- read.csv("reference_tables/incCoding.csv", stringsAsFactors = FALSE)
data3 <- data2%>%
  dplyr::left_join(incCoding,  by = "persIncomeBtax")
data3$Income <- as.numeric(as.character(data3$Income))
medianIncome <- median(data3$Income, na.rm=TRUE)  #dealing with missing data on income using the median
data3$Income <- with(data3, ifelse(is.na(Income), medianIncome, Income))

#AGE: questionable data on age and deleting entries coded "Not applicable"
data3$age <- with(data3, ifelse(age<18 | age=="Not applicable",NA,age))
data3$age <- as.numeric(as.character(data3$age))
data4 <- data3
#ageBin
data4$ageBin <- with(data4, ifelse(age>20&age<=30,"21-30",ifelse(age>30&age<= 40,"31-40",ifelse(age>40&age<=50,"41-50", ifelse(age>50&age<=60,"51-60", ifelse(age>60&age<=70,"61-70", ifelse(age>70&age<=80, "71-80", ifelse(is.na(age),NA, ">80"))))))))
data4$ageBin <- factor(data4$ageBin, levels=c("21-30", "31-40", "41-50", "51-60", "61-70", "71-80",">80"), labels=c("21-30", "31-40", "41-50", "51-60", "61-70", "71-80",">80")) #turning ageBin into factor

#Clean Drain Dry
data4$Clean <- with(data4, ifelse(Clean=="", "Did not report", ifelse(Clean=="half the time","50% of the time",ifelse(Clean=="Every time", "100% of the time",ifelse(Clean=="Never", "0% of the time", Clean)))))

data4 <- subset(data4, select = -c(IPAddress, operateType, persIncomeBtax)) #cleaning up columns no longer needed

#TYPE
data4$type <- with(data4, ifelse(percentPer>0.5, "personal", ifelse(percentCom>0.5, "commercial", "government")))

#wasInElodea
data5 <- data4
data5$wasInElodea <- with(data5, ifelse(wasInElodea=="", NA, wasInElodea))
data5$wasInElodea <- with(data5, 
                          ifelse(stringr::str_detect(wasInElodea,"I DID NOT operate a boat in any of the waterbodies listed"), "No",
                                ifelse(stringr::str_detect(wasInElodea,"I cannot remember"), "Cannot remember", "Yes")))
#freshOutside
data5$freshOutside <- with(data5, ifelse(freshOutside=="Yes","Yes",ifelse(freshOutside=="No","No","Did not report")))
#data5$freshOutside <- factor(data5$freshOutside, levels=c("Yes", "No", "Did not report"), labels=c("Yes", "No", "Did not report"))

data5$Clean <- factor(data5$Clean, levels=c("Did not report","0% of the time","25% of the time","50% of the time","75% of the time","100% of the time"), labels=c("Did not report","0% of the time","25% of the time","50% of the time","75% of the time","100% of the time"))  #need factor for later plot

data5$purchaseType <- with(data5, ifelse(purchaseType=="","Not known",purchaseType)) 
data5$purchaseLoc <- with(data5, ifelse(purchaseLoc=="","Not known",purchaseLoc)) 
# creating subsets for each type of operator
dataCom <- subset(data5, type=="commercial")
dataPers <- subset(data5, type=="personal")
dataGov <- subset(data5, type=="government")
```

Preparing data file for leaflet shiny app
```{r}
dataApp <- data5
dataApp$wasInElodea <- with(dataApp, ifelse(wasInElodea=="Yes","Yes","No"))
dataApp$freshOutside <- with(dataApp, ifelse(freshOutside=="Yes","Yes","No or did not report"))
dataApp$Clean <- with(dataApp, ifelse(Clean=="100% of the time","100% of the time","Less than 100% of the time"))

#Writing data file into appdir for the leaflet shiny app
write.csv(dataApp, "appdir/data/dataApp.csv" )
```



##Sample characteristics
```{r echo=FALSE, results='asis'}
#importing mail return tracking files
busReturns <- read.csv("reference_tables/Business_mailReturn_tracking.csv", stringsAsFactors = F)
privReturns <- read.csv("reference_tables/Private_mailReturn_tracking.csv", stringsAsFactors = F)

#combining business and private dataframe on mail return tracking to form the sample of 5095 contacts
sample <- dplyr::bind_rows(busReturns, privReturns)
#coding NAs
sample$mailReturn <- with(sample, ifelse(is.na(mailReturn), 0, mailReturn))
sample[is.na(sample)]<-0
effsample <- subset(sample, mailReturn==0 )
# calculating % of mail return (non-deliverables), 3.6%, effective sample (N=4914) 
mailReturnRate <- nrow(effsample)/nrow(sample)-1
# map responses (n=372)
responseRate <- nrow(data5)/nrow(effsample) #965/4914 = 19.6%

persResponseRate <- nrow(dataPers)/nrow(privReturns) # 936/4875 = 19.2%  also 8 respondents who were mainly gov operators
comResponseRate <-  nrow(dataCom)/nrow(busReturns) # 21/220 = 9.4%


# Other info about the sample and surveying
otherInfo <- effsample%>%
  summarise(onlineDiffRate = sum(onlineDifficulty)/nrow(effsample),
            noInternetRate = sum(noInternet/nrow(effsample)),
            returnedDollar = sum(returnIncentive),
            removeList = sum(requestsRemove))

# online difficulty 0.5%
# no Internet 0.3%
# 21 returned dollar incentive (answered)
# 2 requested to be removed from list (counted as non-respondent)
```
Of the 5095 mailed envelopes, 3.6% were undeliverable for an effective sample size of 4914. Of these the survey received 965 responses for a response rate of 19.6%. The effective sample contained 220 business contacts which resulted in 21 business responses. Some respondents (0.5%) reported to have online difficulties with mapping their boat tracks and 0.3% reported to not have Internet. 21 respondents returned the $1 bill by mailing the bill back to the research team. Two respondents requested to be taken off the mailing list for the reminder mailings.  

##Representativeness of the sample
We conducted an independent t-test to see how representative respondents to the survey were in comparison to the general Alaska population. Ideally we had hoped to compare the sample to the sample frame (population) of registered boaters but there is no numeric variable that could be of use to conduct such a comparison. The only available metric to conduct this comparison is personal income reported in the survey which can be compared to the personal income (PINCP variable) in the PUMS dataset for Alaska associated with the 2017 American Community Survey. As to be expected, the result of the t-test shows dissimilarity in samples. The median income is much higher in the sample compared to the observed population median. The Figure below shows the sample distribution of income in red while the ACS income distribution is shown in black. 
```{r}
PUMS <- read.csv("reference_tables/PUMS.csv", stringsAsFactors = F)
#F-test if sample variance and ACS variance are the same
var.test(data6$Income, PUMS$PINCP, alternative = "two.sided")  # using PINCP - total person's income closely related to "personal income" the survey asked.  F-test result: variances are not equal, so we need to conduct independent t-test assuming unequal variances.
d1 <- density(data6$Income, na.rm=T)
d2 <- density(PUMS$PINCP, na.rm=T)
plot(d1, col="red", main="Income distributions comparing sample to population",xlab="2017 personal income")
lines(d2)
#R assumes unequal variances by default, which is the Welch two-sample t-test
t.test(data6$Income, PUMS$PINCP, paired=FALSE)
```


##Respondent locations
The  Qualtrics online survey software collects respondent information such as IP addresses which can be used to determine the location the respondent was in when answering the survey. If the respondent completed the survey using the Qualtrics Offline App on a GPS-enabled device, this data will be an accurate representation of the respondent’s location. For all other respondents, the location is an approximation determined by comparing the participant’s IP address to a location database. Inside the United States, this data is typically accurate to the city level.

Figure: Map showing clustered locations for the 965 respondents
```{r}
#creating table of unique respondents and locations 
locations <- data4 %>% 
  distinct(responseID, LocationLatitude, LocationLongitude) %>% 
  drop_na()

data.table::data.table(locations)

leaflet(locations)%>%
    addTiles()%>%
    addProviderTiles(providers$Esri.WorldTopoMap)%>%
    addMarkers(lat = ~LocationLatitude, lng = ~LocationLongitude, 
    clusterOptions = markerClusterOptions(), popup = ~responseID)
```

Figure: Map showing the locations for the 965 respondents 
```{r}
library(leaflet)
leaflet(locations)%>%
    addTiles()%>%
    addProviderTiles(providers$Esri.WorldTopoMap)%>%
    addMarkers(lat = ~LocationLatitude, lng = ~LocationLongitude, popup = ~responseID)
```     

Figure: Map showing locations for 324 respondents with mapping answers
```{r echo=FALSE}
maprespondents <- features4 %>%
  distinct(responseID, LocationLatitude, LocationLongitude) %>% 
  drop_na()

data.table::data.table(maprespondents)

leaflet(maprespondents)%>%
    addTiles()%>%
    addProviderTiles(providers$Esri.WorldTopoMap)%>%
    addMarkers(lat = ~LocationLatitude, lng = ~LocationLongitude , 
    clusterOptions = markerClusterOptions(), popup = ~responseID)
```

Table: Respondent count by user group of 1) solely marine water, 2) solely fresh water, 3) both, and 4) boaters who did not boat in 2018
```{r}
#Number of boaters in each user group who responded
data5$operateFresh1 <- with(data5, ifelse(operateFresh=="Yes"&operateMarine=="No",1,0))
data5$operateMarine1 <- with(data5, ifelse(operateMarine=="Yes"&operateFresh=="No",1,0)) #no data collection for marine boaters
data5$didNotOperate1 <- with(data5, ifelse(operateMarine=="No"&operateFresh=="No",1,0))
data5$operateBoth1 <- with(data5, ifelse(operateMarine=="Yes"&operateFresh=="Yes", 1,0)) #consequently both, marine and fresh water routes were collected
data5$User <- with(data5, ifelse(operateFresh1==1,"Only freshwater",ifelse(operateMarine1==1,"Only marine",ifelse(operateBoth1==1,"Fresh and marine","Did not operate"))))

userGroups <- data5%>%
  summarise(FreshOnly = round((sum(operateFresh1)/nrow(data5))*100,digits=0),
            MarineOnly = round((sum(operateMarine1)/nrow(data5))*100,digits=0),
            Both = round((sum(operateBoth1)/nrow(data5))*100,digits=0),
            DidNot = round((sum(didNotOperate1)/nrow(data5))*100,digits=0),
            FreshOnlyCount = sum(operateFresh1),
            BothCount = sum(operateBoth1))

knitr::kable(userGroups, caption = "Percent of respondents and where they operated boats in 2018, n=965")%>%
  kableExtra::kable_styling()
```

Tables: Number of respondents with map answers, duration of response (if map answered), by user group
```{r}
data5 <- data5%>%
  left_join(maprespondents, by = "responseID")
data5$mapres <- with(data5, ifelse(!is.na(LocationLatitude.y),1,0))
data6 <- select(data5, -c("LocationLatitude.y","LocationLongitude.y"))

mapResponse <- subset(data6, mapres==1)
mapNonRes <- subset(data6, mapres==0)

#Summary of respondents with mapping response
mapResSummary <- mapResponse%>%
  group_by(User)%>%
    summarise(median.min = round(median(Duration.s)/60, digits = 1),
            count = n(),
            elodeaCount = sum(wasInElodea=="Yes", na.rm = TRUE))
knitr::kable(mapResSummary, caption = "Summary of respondents with mapping response, n=322")%>%
  kableExtra::kable_styling()
#Summary of respondents without mapping response
non.mapResSummary <- mapNonRes%>%
  group_by(User)%>%
  summarise(median.min = round(median(Duration.s)/60, digits = 1),
            count = n(),
            elodeaCount = sum(wasInElodea=="Yes", na.rm = TRUE))
knitr::kable(non.mapResSummary, caption = "Summary of respondents without mapping response, n=643")%>%
  kableExtra::kable_styling()
```

Table: Respondent count for boat owners who took their boat into elodea infested waterbodies in 2018, n=51 
```{r}
elodeaLocFreq <- features4%>%
  distinct(responseID, .keep_all=T)%>%  #Note, important to set the dot in front of keep_all!!!
  filter(wasInElodea=="Yes")%>%
  select(responseID, elodeaLoc1, elodeaLoc2, elodeaLoc3)%>% 
  gather(type, value=location, elodeaLoc1, elodeaLoc2, elodeaLoc3)%>%
  filter(location!="I cannot remember"&location!=""&location!="Campbell Lake")%>%  #taking out blanks and miscoding
  group_by(location)%>%
  summarise(count=n())

knitr::kable(elodeaLocFreq, caption = "Count of respondents visiting waterbodies with known elodea infestations in 2018, n=51")%>%
  kableExtra::kable_styling()
```

Table: Descriptive statistics related to personal operators who gave mapping answers, n=324
```{r}
dataSummary <- features4 %>%
  distinct(responseID, .keep_all=T)%>%
  filter(type=="personal")%>%
  select(Income, age, pax, routeCount,totalKmPre,hCost, YrCostPre, totalTripsPre, totalTripsPost) # select variables to summarise

tidySummary <- summarytools::descr(dataSummary, stats = c("mean", "sd", "cv","min","q1","med","q3","max","pct.valid"))
Statistic <- row.names(tidySummary) # extracting row names from the matrix above
tidySummary<- data.frame(tidySummary)%>% 
  mutate_at(1:9, funs(round(.,0)))
tidySummary <-cbind(Statistic, tidySummary) #adding variable names back in

names(tidySummary)[names(tidySummary)=="age"] <- "Age"#renaming misspelled column
names(tidySummary)[names(tidySummary)=="hCost"] <- "Operating cost $/h"
names(tidySummary)[names(tidySummary)=="YrCostPre"] <- "Operating cost $/year"
names(tidySummary)[names(tidySummary)=="totalTripsPre"] <- "Trips/year"
names(tidySummary)[names(tidySummary)=="totalKmPre"] <- "Km/year"
names(tidySummary)[names(tidySummary)=="totalTripsPost"] <- "Trips/year contingent"
names(tidySummary)[names(tidySummary)=="pax"] <- "Passengers"
names(tidySummary)[names(tidySummary)=="routeCount"] <- "Number of unique routes"

knitr::kable(tidySummary, caption = "Summary statistics for personal operators with mapping answers, n=324")%>%
  kableExtra::kable_styling()
```

Table: Descriptive statistics related to all respondents' AIS concern for specific sectors , n=372
```{r}
concSummary <- features4 %>%
  distinct(responseID, .keep_all=T)%>%
  select(sportfishConc,boatingSafetyConc,recValueConc,businessConc,realEstateConc,subsistConc,biodivConc,commFishConc) # select variables to summarise

tidyCon <- summarytools::descr(concSummary, stats = c("min","med","max","pct.valid"))
Statistic <- row.names(tidyCon) # extracting row names from the matrix above
tidyCon<- data.frame(tidyCon)%>% 
  mutate_at(1:8, funs(round(.,0)))
tidyCon <-cbind(Statistic, tidyCon) #adding variable names back in

names(tidyCon)[names(tidyCon)=="sportfishConc"] <- "Sport fishing"#renaming column
names(tidyCon)[names(tidyCon)=="boatingSafetyConc"] <- "Boating safety"
names(tidyCon)[names(tidyCon)=="recValueConc"] <- "Recreation value"
names(tidyCon)[names(tidyCon)=="businessConc"] <- "Businesses"
names(tidyCon)[names(tidyCon)=="realEstateConc"] <- "Real estate values"
names(tidyCon)[names(tidyCon)=="subsistConc"] <- "Subsistence"
names(tidyCon)[names(tidyCon)=="biodivConc"] <- "Biodiversity"
names(tidyCon)[names(tidyCon)=="commFishConc"] <- "Commercial fishing"

knitr::kable(tidyCon, caption = "Summary statistics for all rspondents' AIS concerns in specific sectors, n=324")%>%
  kableExtra::kable_styling()
```

##Alaska boat owners' boating and boat purchases occurring outside of Alaska
Table: Boat owner count for owners who took their boat out of state before boating in Alaska in 2018
```{r}
#Total distance, total trips, total respondents, and weighted average distance per trip by age group in the personal user group
outsidersSummary <- features4%>%
  group_by(locOutside, freshOutside)%>%
  summarise(count = n_distinct(responseID),
            percentOfN = count/333)%>%
  na.omit()

knitr::kable(outsidersSummary, caption = "Percentage of boat owners boating outside of Alaska in 2018")%>%
  kableExtra::kable_styling()
```

Table: Boat owner characteristics who boated outside Alaska before boating in Alaska
```{r}
outsideSummary <- features4 %>%
  distinct(responseID, .keep_all=T)%>%
  filter(freshOutside=="Yes")%>%
  select(Income, age, pax, routeCount,totalKmPre,hCost, YrCostPre, totalTripsPre, totalTripsPost) # select variables to summarise

Summary <- summarytools::descr(outsideSummary, stats = c("mean", "sd", "cv","min","q1","med","q3","max","pct.valid"))
Statistic <- row.names(Summary) # extracting row names from the matrix above
Summary<- data.frame(Summary)%>% 
  mutate_at(1:9, funs(round(.,0)))
Summary <-cbind(Statistic, Summary) #adding variable names back in

names(Summary)[names(Summary)=="age"] <- "Age"#renaming misspelled column
names(Summary)[names(Summary)=="hCost"] <- "Operating cost $/h"
names(Summary)[names(Summary)=="YrCostPre"] <- "Operating cost $/year"
names(Summary)[names(Summary)=="totalTripsPre"] <- "Trips/year"
names(Summary)[names(Summary)=="totalKmPre"] <- "Km/year"
names(Summary)[names(Summary)=="totalTripsPost"] <- "Trips/year contingent"
names(Summary)[names(Summary)=="pax"] <- "Passengers"
names(Summary)[names(Summary)=="routeCount"] <- "Number of unique routes"

knitr::kable(Summary, caption = "Summary statistics for operators who took their boats outside first in 2018, n=4")%>%
  kableExtra::kable_styling()
```


Table: Outside purchases of boats
```{r}
purOutside <- data6%>%
  filter(purchaseOutside=="Yes")%>%
  group_by(purchaseType,purchaseYear)%>%
  summarise(count = n())

h1 <- purOutside
h2 <- expand.grid(purchaseType=unique(h1$purchaseType), purchaseYear=1982:2018)
h3 <- merge(h2,h1, by.x=c("purchaseType", "purchaseYear"),by.y=c("purchaseType", "purchaseYear"), all.x = TRUE)
h3[is.na(h3)]<-0

h3plot <- ggplot(h3, aes(x = purchaseYear, y = count, fill = purchaseType)) +
  geom_bar(stat = 'identity', aes(fill = purchaseType)) +
  labs(x="Year",y="Count", title="Historical purchases of boats from outside Alaska, n=459") + 
  scale_fill_discrete(name = "Purchase type") + theme_bw()
h3plot
```

Map: States where boats were purchased by type of purchase
```{r}
purLoc <- data6%>%
  filter(purchaseOutside=="Yes")%>%
  group_by(purchaseLoc,purchaseType)%>%
  summarise(n = n())%>%
  mutate(percent = n/sum(n))
#eliminating unknown locations
purLoc <- purLoc[!purLoc$purchaseLoc=="Not known",]

purLoc <- subset(purLoc, select = -n) #cleaning up columns no longer needed
  purLoc<-purLoc%>%
  spread(key=purchaseType, value=percent)

#adding georeferences
locs_center <- read.csv("reference_tables/locs_center.csv", stringsAsFactors = F)
purLocGeo <- purLoc%>%
  left_join(locs_center, by=c("purchaseLoc"="state"))
purLocGeo[is.na(purLocGeo)] <- 0

#Minichart map for proportion of used new boats being bought
library(leaflet)

basemap <- leaflet(width = "100%", height = "650px") %>%
  addTiles()%>%
  addProviderTiles(providers$OpenStreetMap.Mapnik)

colors <- c("#1b9e77","#d95f02","#7570b3")

basemap %>%
  addMinicharts(
    purLocGeo$long, purLocGeo$lat,
    type = "pie",
    chartdata = purLocGeo[, c("New","Not known","Used")], 
    colorPalette = colors, 
    width = 50 , transitionTime = 0)
```




##Survey data analysis
Table: Total distance, total trips, total respondents, and weighted average distance per trip by age group in the personal user group
```{r}
ageBinSummary <- featuresPers%>%
  subset(totalKmPre < 2000)%>%
  group_by(ageBin)%>%
  summarise(count = n_distinct(responseID),
            sumTotalKm = round(sum(totalKmPre),digits =0),
            sumTrips  = sum(freqPre),
            wghtAvgDis = round(sumTotalKm/sumTrips, digits=1))
knitr::kable(ageBinSummary, caption = "Total distance, trips, respondent count, and weighted average distance per trip by age group")%>%
  kableExtra::kable_styling()
```

Table: Personal trip characteristics by age group
```{r}
#Total distance, total trips, total respondents, and weighted average distance per trip by age group in the personal user group
ageBinSummary <- featuresPers%>%
  subset(totalKmPre < 2000)%>%
  group_by(ageBin)%>%
  summarise(count = n_distinct(responseID),
            sumTotalKm = round(sum(totalKmPre),digits =0),
            sumTrips  = sum(freqPre),
            wghtAvgDis = round(sumTotalKm/sumTrips, digits=1))
knitr::kable(ageBinSummary, caption = "Total distance, trips, respondent count, and weighted average distance per trip by age group")%>%
  kableExtra::kable_styling()
```

Table: Statistics of main purpose of route showing main purpose of route by mean ranking
```{r}
StatsByPur <- features4 %>%
  dplyr::group_by(primaryPur)%>%
  summarise(routeCount = n(),
            percentAllRoutes = round((n()/nrow(features4))*100,digits=0),
            meanRank = round(mean(rank),digits=0),
            meanKm = round(mean(km),digits=0),
            minKm = round(min(km),digits=0),
            maxKm= round(max(km), digits=0))%>%
  arrange(desc(routeCount))%>%
  na.omit()

names(StatsByPur)[names(StatsByPur)=="primaryPur"] <- "Primary purpose"
names(StatsByPur)[names(StatsByPur)=="routeCount"] <- "Route count"#renaming column
names(StatsByPur)[names(StatsByPur)=="percentAllRoutes"] <- "% of all routes"
names(StatsByPur)[names(StatsByPur)=="meanRank"] <- "Mean rank"
names(StatsByPur)[names(StatsByPur)=="meanKm"] <- "Average route in km"
names(StatsByPur)[names(StatsByPur)=="minKm"] <- "Shortest route in km"
names(StatsByPur)[names(StatsByPur)=="maxKm"] <- "Longest route in km"

knitr::kable(StatsByPur, caption = "Summary statistics for the primary purpose of routes, n=324")%>%
  kableExtra::kable_styling()
```

##Operating cost
Tables: Annual operating cost followed by operating cost per hour by operator type 
```{r}
resSum <- features4%>%
  group_by(responseID, type)%>%
  summarise(YrTravelCost = sum(YrCostPre))

#Calculating descriptive stats for annual operating cost by operator type
TCByType <- resSum%>%
  group_by(type)%>%
  summarise(count = n(),
            mean = round(mean(YrTravelCost,na.rm = TRUE),digits=1),
            median = round(median(YrTravelCost,na.rm = TRUE),digits=1),
            sd = round(sd(YrTravelCost,na.rm = TRUE),digits=1),
            max = round(max(YrTravelCost,na.rm = TRUE),digits=1),
            min = round(min(YrTravelCost,na.rm = TRUE),digits=1),
            cv = round(sd/mean, digits=1))
knitr::kable(TCByType, caption = "Estimated annual operating cost by operator type")%>%
  kableExtra::kable_styling()

hCByType <- features4%>%
  group_by(type)%>%
  summarise(count = n(),
            mean = round(mean(hCost,na.rm = TRUE),digits=1),
            median = round(median(hCost,na.rm = TRUE),digits=1),
            sd = round(sd(hCost,na.rm = TRUE),digits=1),
            max = round(max(hCost,na.rm = TRUE),digits=1),
            min = round(min(hCost,na.rm = TRUE),digits=1),
            cv = round(sd/mean, digits=1))

knitr::kable(hCByType, caption = "Stated hourly operating cost by operator type")%>%
  kableExtra::kable_styling()
```




##Clean Drain Dry best practices
Figure: Stated percentage of time respondent followed clean-drain-dry procedures to minimize transmission of aquatic invasive species by age and annual distance
```{r}
#dealing with 50,000 total km of boat travel outlier
pers3 <- featuresPers%>%
  subset(totalKmPre < 2000 &
         Clean != "Did not report")

#colors for plot
theme <- c('#a63603','#e6550d','#fd8d3c','#fdae6b','#fdd0a2','#feedde')
#plot
ggplot(pers3, aes(age, totalKmPre, colour = Clean)) + 
  geom_point() +
  geom_hline(aes(yintercept = 0)) + ylim(0,2000) + ylab("km / year")+
  labs(colour = "CleanDrainDry") + scale_color_manual(values=theme) 
```

Figure: Stated percentage of time respondent followed clean-drain-dry procedure to minimize transmission of aquatic invasive species by age group (n=312). 
```{r}
persAgeBinClean <- dataPers %>%
    filter(!is.na(ageBin)) %>%
    group_by(ageBin, Clean) %>%
    summarise(n = n())
#965 total respondents, 302 reported age and Clean, 639 did not report Clean, 653 not reporting age
#plotting proportions of clean drain dry proportions by age bin
library(ggplot2)
plot1 <- ggplot(persAgeBinClean, aes(x = ageBin, y = n, fill = Clean)) +
  geom_bar(stat = 'identity', position = 'fill', aes(fill = Clean)) +
  labs(x="age group",y="proportion") + 
  scale_fill_discrete(name = "Clean Drain Dry") + theme_bw() 

plot2 <- ggplot(persAgeBinClean, aes(x = ageBin, y = n, fill = Clean, label = n)) +
  geom_bar(stat = 'identity', aes(fill = Clean)) +
  labs(x="age group",y="count") + 
  scale_fill_discrete(name = "Clean Drain Dry") + 
  geom_text(aes(y = n, label = n),size = 3, check_overlap = TRUE, position = position_stack(vjust = 0.5)) + 
  theme_bw()

ggpubr::ggarrange(plot1, plot2, nrow=1, ncol=2, common.legend = TRUE, legend = "right")
```

Figure: Map of respondents and where they reside color coded by whether repondents gave mapping answers or did not boat (n=848) 
```{r}
vars <- c("mapres","responseID","LocationLongitude.x","LocationLatitude.x","didNotOperate1")
respondentLocations <- data6[vars]
#eliminting respondents where we could not get location information
respondentLocations <- subset(respondentLocations, !is.na(LocationLongitude.x))
respondentLocations$marker <- with(respondentLocations, ifelse(mapres==1, 1, ifelse(mapres==0&didNotOperate1==0,0,2)))

getColor <- function(respondentLocations) {
  sapply(respondentLocations$marker, function(marker) {
  if(marker==1) {
    "green"
  } else if(marker==0) {
    "red"
  } else {
    "blue"
  } })
}

icons <- awesomeIcons(
  icon = 'ios-close',
  iconColor = 'black',
  library = 'ion',
  markerColor = getColor(respondentLocations)
)

leaflet(respondentLocations)%>%
    addTiles()%>%
    addProviderTiles(providers$Esri.WorldTopoMap)%>%
    addAwesomeMarkers(~LocationLongitude.x, ~LocationLatitude.x, icon=icons, label=~as.character(mapres))%>%
    addLegend("topright", colors = c("green","red", "blue"), labels = c("gave map response","no map response","did not boat"))
```

##Elodea transmission risk map
Figure: Respondents' (n=324) boat routes and locations of known elodea occurrence
```{r}
#importing elodea presence data, and boating tracks
library(geojsonio)
library(leaflet)
elodea <- read.csv("appdir/data/elodea.csv", stringsAsFactors = FALSE)
elodeaTRUE <- subset(elodea, presence=="TRUE")
Alltracks <- geojson_read("appdir/data/map.geojson",what = "sp")
NotClean <- geojson_read("appdir/data/NotClean.geojson",what = "sp")

#Interactive layer display
clickMap1 <- leaflet(elodeaTRUE, width = "80%", height = 500) %>%
  setView(-147.3, 62.0, zoom = 5)%>%
  #Base groups
  addTiles(group = "OpenStreetMap") %>%
  addProviderTiles(providers$Esri.WorldTopoMap, group = "Esri Topo")%>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Esri Image")%>%
  #Overlay groups
  addCircleMarkers(lat = ~latitude, lng = ~longitude, ~log(infested_area)*10, stroke = F, 
                   color="#d95f02", weight = 5, group="Elodea infestation" )%>%
  addPolylines(data=Alltracks, stroke = TRUE, color = "#1b9e77", weight = 2,
               opacity = 1.0, fill = FALSE, group = "All boat tracks (green)")%>%
  addPolylines(data=NotClean, stroke = TRUE, color = "#7570b3", weight = 2,
               opacity = 1.0, fill = FALSE, group = "No clean-drain-dry (purple)")%>%
  #Layers control
  addLayersControl(
    baseGroups = c("Esri Topo", "Esri Image","OpenStreetMap"),
    overlayGroups = c("Elodea infestation", "All boat tracks","No clean drain dry"),
    options = layersControlOptions(collapsed = FALSE)
  )
clickMap1
```




