---
title: "Analysis of responses from a statewide survey with registered boat owners in Alaska, fall 2018"
author: "Tobias Schwoerer"
date: "November 21, 2018"
output: html_document
editor_options: 
  chunk_output_type: inline
---

# Importing the data files, visualizing all respondent boat tracks 

```{r}
library(jsonlite)
library(geojson)
library(leaflet)
library(dplyr)
library(tidyr)
library(Hmisc)
library(ggplot2)
library(RColorBrewer)

#cleaned up features (tracks the respondents entered into the mapping tool)
#map <- geojson_read("D:/Dropbox/DATA/2018_boat_survey/map.geojson",what = "sp")
#map <- geojson_read("C:/Users/Toby/Dropbox/DATA/2018_boat_survey/map.geojson", what = "sp")

#features <- read.csv("D:/Dropbox/DATA/2018_boat_survey/features.csv", stringsAsFactors = FALSE)
features <- read.csv("C:/Users/Toby/Dropbox/DATA/2018_boat_survey/features.csv", stringsAsFactors = FALSE)

#renaming misspelled column
names(features)[names(features)=="LenghtKM"] <- "km"


freqCoding <- read.csv("reference_tables/freqCoding.csv", stringsAsFactors = FALSE)

#cleaned data associated with the survey answers of each respondent
#data <- read.csv("D:/Dropbox/DATA/2018_boat_survey/data.csv", stringsAsFactors = FALSE)
data <- read.csv("C:/Users/Toby/Dropbox/DATA/2018_boat_survey/data.csv", stringsAsFactors = FALSE)

# the joined data and map output, joined these files using the python script
#mapdata <- geojson_read("D:/Dropbox/DATA/2018_boat_survey/output.geojson",what = "sp")
#mapdata <- geojson_read("C:/Users/Toby/Dropbox/DATA/2018_boat_survey/output.geojson",what = "sp")

#creating leaflet map showing all routes
#leaflet(mapdata) %>%
#  addTiles() %>%
#  addPolylines() %>%
#  addProviderTiles(providers$Stamen.TerrainBackground)
```

#Calculating trip frequencies pre and post hypothetical elodea invasions of waterways used 
```{r}
#calculating pre elodea trip frequencies
features2 <- features%>%
  left_join(freqCoding,  by = c("trackFrequ"="Coding"))
features2$freqPre <- rowMeans(features2[c("Lower","Upper")], na.rm=TRUE)
#Renaming columns 
names(features2)[names(features2)=="Lower"] <- "LowerPre"
names(features2)[names(features2)=="Upper"] <- "UpperPre"

#Calculating post elodea trip frequencies
features3 <- features2 %>%
  left_join(freqCoding,  by = c("trackFre_1"="Coding"))
features3$freqPost <- rowMeans(features3[c("Lower","Upper")], na.rm=TRUE)
#Renaming columns 
names(features3)[names(features3)=="Lower"] <- "LowerPost"
names(features3)[names(features3)=="Upper"] <- "UpperPost"
```

#Outliers and missing data for features
#Calculating annual cost of boating per route assumes average travel speed of 10 miles/h equal to 16km/h
```{r}
#calculating mean cost from the high and low cost per hour respondents reported
#last argument excludes missing values
features3$hCost <- rowMeans(features3[c("hCostHigh","hCostLow")], na.rm=TRUE)

#dealing with outlier on cost
features3$hCost[features3$hCost==20000] <- NA
hist(features3$hCost, breaks=20)

#dealing with outliers on km, setting frequency of trips pre if Up to 4 setting it to 1 if km >300
features3$freqPre <- with(features3, ifelse(km>300 & trackFrequ=="Up to 4",1,freqPre))

#dealing with ranks that are 0
features3$rank <- with(features3, ifelse(rank==0, 1, rank)) 

#dealing with missing values by first calculating the median cost per h then inserting missing values with the median
medianCost <- median(features3$hCost, na.rm=TRUE)
features3$hCost <- with(features3, ifelse(is.na(hCost), medianCost, hCost))

#calculating the total km per route then annual cost per route
features3$totalKmPre <- with(features3, km * freqPre)
features3$totalKmPost <- with(features3, km * freqPost)
features3$YrCostPre <- with(features3, totalKmPre * hCost)
features3$YrCostPost <- with(features3, totalKmPost * hCost)
```

#cleaning up features further, missing data etc. 
```{r}
#getting rid of % sign in character string, then converting character to numeric, doing for percentCom and percentGov
features4 <- features3 
features4$percentCom <- gsub("[[:punct:]]", " ", features4$percentCom)
features4$percentCom <- as.numeric(as.character(features4$percentCom))
features4$percentCom <- with(features4, ifelse(is.na(percentCom),0,percentCom/100))
features4$percentGov <- gsub("[[:punct:]]", " ", features4$percentGov)
features4$percentGov <- as.numeric(as.character(features4$percentGov))
features4$percentGov <- with(features4, ifelse(is.na(percentGov),0,percentGov/100))
features4$percentPer <- with(features4, 1-percentGov-percentCom)

#correcting wrong respondent entry on percentGov
features4$percentGov <- with(features4, ifelse(percentPer<0,percentGov-percentCom,percentGov))
features4$percentPer <- with(features4, 1-percentGov-percentCom)

#renaming columns
names(features4)[names(features4)=="CleanDrainDry"] <- "Clean"

#turning character to numeric for income variable
incCoding <- read.csv("reference_tables/incCoding.csv", stringsAsFactors = FALSE)
features4 <- features4%>%
  dplyr::left_join(incCoding,  by = "persIncomeBtax")
features4$Income <- as.numeric(as.character(features4$Income))

#dealing with missing data on income
#dealing with missing values by first calculating the median income then inserting missing values with the median
medianIncome <- median(features4$Income, na.rm=TRUE)
features4$Income <- with(features4, ifelse(is.na(Income), medianIncome, Income))

#dealing with questionable data on age and deleting entries coded "Not applicable"
features4$age <- with(features4, ifelse(age==9 | age=="Not applicable",NA,age))
features4$age <- as.numeric(as.character(features4$age))

#coding respondents who did not respond to Clean Drain Dry question
features4$Clean <- with(features4, ifelse(Clean=="", "Did not report", ifelse(Clean=="half the time","50% of the itme",ifelse(Clean=="Every time", "100% of the time",ifelse(Clean=="Never", "0% of the time", Clean)))))

#coding respondents who did not respond to wasInElodea
library(stringr)
features4$wasInElodea<-str_replace_all(features$wasInElodea,"\\s","_")
features4$wasInElodea <- with(features4, ifelse(wasInElodea=="","Did not report",wasInElodea))
features4$wasInElodea <- with(features4, ifelse(wasInElodea=="I_cannot_remember","Cannot remember",wasInElodea))
write.csv(featuresApp, "C:/Users/Toby/Dropbox/DATA/2018_boat_survey/featuresApp.csv" )

# creating subsets for each type of operator
com <- subset(features5, type=="commercial")
pers <- subset(features5, type=="personal")
gov <- subset(features5, type=="government")
```

# hourly and yearly operating cost statistics (distributions)
```{r}
resSum <- features4%>%
  group_by(responseID, type)%>%
  summarise(YrTravelCost = sum(YrCostPre))

#Calculating descriptive stats for annual operating cost by operator type
TCByType <- resSum%>%
  group_by(type)%>%
  summarise(mean = mean(YrTravelCost),
            median = median(YrTravelCost),
            sd = sd(YrTravelCost),
            max = max(YrTravelCost),
            min = min(YrTravelCost),
            cv = sd/mean)

hCByType <- features5%>%
  group_by(type)%>%
  summarise(mean = mean(hCost),
            median = median(hCost),
            sd = sd(hCost),
            max = max(hCost),
            min = min(hCost),
            cv = sd/mean)
```

#Plotting distance by personal operators by age  NEEDS WEIGHTING
```{r}
#dealing with 50,000 total km of boat travel outlier
kmDistribution <- pers%>%
  subset(totalKmPre < 2000)%>%
  group_by(ageBin)%>%
  summarise(ageBinKm = sum(totalKmPre))%>%
  na.omit()

ggplot(kmDistribution, aes(x = ageBin, y = ageBinKm)) +
  geom_bar(stat = "identity")+ theme_bw() + labs(title="2018 Total Distance of Personal Boat Trips by Age Group",x="Age Group",y="km", caption = "Note, does not show almost 28,000km of personal boaters who did not report their age.")

```

#Plotting distance compliance with clean drain dry by age using the features data
```{r}
#dealing with 50,000 total km of boat travel outlier
pers2 <- pers%>%
  subset(totalKmPre < 2000)
pers3 <- pers%>%
  subset(totalKmPre < 2000 &
         Clean != "Did not report")

#colors for plot
theme <- c('#a63603','#e6550d','#fd8d3c','#fdae6b','#fdd0a2','#feedde')
#plot
t <- ggplot(pers3, aes(age, totalKmPre, colour = Clean)) + geom_point() +
  geom_hline(aes(yintercept = 0)) + ylim(0,2000) 
t + labs(colour = "CleanDrainDry") +scale_color_manual(values=theme) +ylab("km / year")
```


#Clean Drain Dry behavior by age bin and operator type
# separating commercial from personal in data.csv
```{r}
#getting rid of % sign in character string, then converting character to numeric, doing for percentCom and percentGov
data2 <- data
data2$percentCom <- gsub("[[:punct:]]", " ", data2$percentCom)
data2$percentCom <- as.numeric(as.character(data2$percentCom))
data2$percentCom <- with(data2, ifelse(is.na(percentCom),0,percentCom/100))
data2$percentGov <- gsub("[[:punct:]]", " ", data2$percentGov)
data2$percentGov <- as.numeric(as.character(data2$percentGov))
data2$percentGov <- with(data2, ifelse(is.na(percentGov),0,percentGov/100))
data2$percentPer <- with(data2, 1-percentGov-percentCom)

#correcting wrong respondent entry on percentGov
data2$percentGov <- with(data2,ifelse(percentPer<0,percentGov-percentCom,percentGov))
data2$percentPer <- with(data2, 1-percentGov-percentCom)

#renaming columns
names(data2)[names(data2)=="CleanDrainDry"] <- "Clean"

#turning character to numeric for income variable
incCoding <- read.csv("reference_tables/incCoding.csv", stringsAsFactors = FALSE)
data3 <- data2%>%
  dplyr::left_join(incCoding,  by = "persIncomeBtax")
data3$Income <- as.numeric(as.character(data3$Income))

#dealing with missing data on income
#dealing with missing values by first calculating the median income then inserting missing values with the median
medianIncome <- median(data3$Income, na.rm=TRUE)
data3$Income <- with(data3, ifelse(is.na(Income), medianIncome, Income))

#dealing with questionable data on age and deleting entries coded "Not applicable"
data3$age <- with(data3, ifelse(age<18 | age=="Not applicable",NA,age))
data3$age <- as.numeric(as.character(data3$age))

#coding respondents who did not respond to Clean Drain Dry question
data3$Clean <- with(data3, ifelse(Clean=="", "Did not report", ifelse(Clean=="half the time","50% of the time",ifelse(Clean=="Every time", "100% of the time",ifelse(Clean=="Never", "0% of the time", Clean)))))

#turning character vectors into factors for Clean variable, for later when we need a nice legend for the plot
data3$Clean <- factor(data3$Clean, levels=c("Did not report","0% of the time","25% of the time","50% of the time","75% of the time","100% of the time"), labels=c("Did not report","0% of the time","25% of the time","50% of the time","75% of the time","100% of the time"))

#cleaning up columns no longer needed
data4 <- subset(data3, select = -c(IPAddress, operateType, persIncomeBtax))

#Creating bins for age
data4$ageBin <- with(data4, ifelse(age>20&age<=30,"21-30",ifelse(age>30&age<= 40,"31-40",ifelse(age>40&age<=50,"41-50", ifelse(age>50&age<=60,"51-60", ifelse(age>60&age<=70,"61-70", ifelse(age>70&age<=80, "71-80", "81 and over")))))))
#turning ageBin into factor
data4$ageBin <- factor(data4$ageBin, levels=c("21-30", "31-40", "41-50", "51-60", "61-70", "71-80","81 and over"), labels=c("21-30", "31-40", "41-50", "51-60", "61-70", "71-80","81 and over"))

#creating subsets by operator type
#creating variable for personal and commercial operators
data4$type <- with(data4, ifelse(percentPer>0.5, "personal", ifelse(percentCom>0.5, "commercial", "government")))

# creating subsets for each type of operator
datacom <- subset(data4, type=="commercial")
datapers <- subset(data4, type=="personal")
datagov <- subset(data4, type=="government")
```

#Plotting compliance with clean drain dry by age bin using the data.csv file with 965 observations not only the respondents who used the mapping tool
```{r}
persAgeBinClean <- datapers %>%
    filter(!is.na(ageBin)) %>%
    group_by(ageBin, Clean) %>%
    summarise(n = n())
sum(persAgeBinClean$n)
#965 total respondents, 302 reported age and Clean, 639 did not report Clean, 653 not reporting age
#plotting proportions of clean drain dry proportions by age bin
library(ggplot2)
ggplot(persAgeBinClean, aes(x = ageBin, y = n, fill = Clean)) +
       geom_bar(stat = 'identity', position = 'fill', aes(fill = Clean)) +labs(x="Age Group",y="proportion", caption = "302 respondents reported age and extent of clean drain dry.") + scale_fill_discrete(name = "Clean Drain Dry") 

#trying to get n numbers on the top of each bar

ggplot(persAgeBinClean, aes(x = ageBin, y = n, fill = Clean, label = n)) +
       geom_bar(stat = 'identity',position = 'fill', aes(fill = Clean)) +labs(x="Age Group",y="proportion", caption = "302 respondents reported age and extent of clean drain dry.") + scale_fill_discrete(name = "Clean Drain Dry") + geom_text(size = 3)

```

#Creating map of response locations shown for all 965 respondents
```{r}
#creating table of unique respondents and locations 
locations <- data4 %>% 
  distinct(responseID, LocationLatitude, LocationLongitude) %>% 
  drop_na()

data.table::data.table(locations)

leaflet(locations)%>%
    addTiles()%>%
    addProviderTiles(providers$Esri.WorldTopoMap)%>%
    addMarkers(lat = ~LocationLatitude, lng = ~LocationLongitude, 
    clusterOptions = markerClusterOptions(), popup = ~responseID)

library(leaflet)
leaflet(locations)%>%
    addTiles()%>%
    addProviderTiles(providers$Esri.WorldTopoMap)%>%
    addMarkers(lat = ~LocationLatitude, lng = ~LocationLongitude, popup = ~responseID)

```     

#Creating map of response locations (324 respondents with mapping answers)
```{r echo=FALSE}
maprespondents <- features5 %>%
  distinct(responseID, LocationLatitude, LocationLongitude) %>% 
  drop_na()

data.table::data.table(maprespondents)

leaflet(maprespondents)%>%
    addTiles()%>%
    addProviderTiles(providers$Esri.WorldTopoMap)%>%
    addMarkers(lat = ~LocationLatitude, lng = ~LocationLongitude, 
    clusterOptions = markerClusterOptions(), popup = ~responseID)
```


#plotting features data and map of elodea occurences
```{r}
#importing elodea presence data, and boating tracks
elodea <- read.csv("C:/Users/Toby/Dropbox/DATA/2018_boat_survey/elodea.csv")
elodeaTRUE <- subset(elodea, presence=="TRUE")

mapgeo <- geojson_read("C:/Users/Toby/Dropbox/DATA/2018_boat_survey/map.geojson", what = "sp")

#creating leaflet map of elode occurences
leaflet(elodeaTRUE, width = "75%", height = 500) %>%
  addTiles() %>%
  addProviderTiles(providers$Esri.WorldTopoMap)%>%
  addCircleMarkers( lat = ~latitude, lng = ~longitude, radius=5, stroke = F, 
                   color="red", weight = 5, clusterOptions = markerClusterOptions(spiderLegPolylineOptions = list(weight = 1.5, color = "#222", 
                                                    opacity = 0.5)))
                   
#creating leaflet map of boat tracks
leaflet(mapgeo, width = "75%", height = 500)%>%
  addTiles() %>%
  addProviderTiles(providers$Esri.WorldTopoMap)%>%
  addPolylines(stroke = TRUE, color = "green", weight = 2,
  opacity = 1.0, fill = FALSE, fillColor = "#03F",
  fillOpacity = 0.2)
```

#response rate, effective sample, etc. 
```{r echo=FALSE, results='asis'}
#importing mail return tracking files
busReturns <- read.csv("C:/Users/Toby/Dropbox/DATA/2018_boat_survey/Business_mailReturn_tracking.csv", stringsAsFactors = F)
privReturns <- read.csv("C:/Users/Toby/Dropbox/DATA/2018_boat_survey/Private_mailReturn_tracking.csv", stringsAsFactors = F)

#combining business and private dataframe on mail return tracking to form the sample of 5095 contacts
sample <- dplyr::bind_rows(busReturns, privReturns)
#coding NAs
sample$mailReturn <- with(sample, ifelse(is.na(mailReturn), 0, mailReturn))
sample[is.na(sample)]<-0
effsample <- subset(sample, mailReturn==0 )
# calculating % of mail return (non-deliverables), 3.6%, effective sample (N=4914) 
mailReturnRate <- nrow(effsample)/nrow(sample)-1
# map responses (n=372)
responseRate <- nrow(data4)/nrow(effsample) #965/4914 = 19.6%

persResponseRate <- nrow(datapers)/nrow(privReturns) # 936/4875 = 19.2%  also 8 respondents who were mainly gov operators
comResponseRate <-  nrow(datacom)/nrow(busReturns) # 21/220 = 9.4%

#Number of boaters in each user group who responded
data5 <- data4
data5$operateFresh <- with(data5, ifelse(operateFresh=="Yes",1,0))
data5$operateMarine <- with(data5, ifelse(operateMarine=="Yes",1,0))
data5$didNotOperate <- with(data5, ifelse(operateMarine==0&operateFresh==0,1,0))
data5$operateBoth <- with(data5, ifelse(operateMarine==1&operateFresh==1, 1,0))

userGroups <- data5%>%
  summarise(fresh = sum(operateFresh),
            marine = sum(operateMarine),
            both = sum(operateBoth),
            didNot = sum(didNotOperate))

#Number of respondents with map answers, duration of response (if map answered), by user group
data6 <- data5%>%
  left_join(maprespondents, by = "responseID")
data6$mapres <- with(data6, ifelse(!is.na(LocationLatitude.y),1,0))
data6 <- select(data6, -c("LocationLatitude.y","LocationLongitude.y"))

## respondents who were in elodea, new variable, Elodea with 1- yes, 0-no, I cannot remember coded as NA
data6$Elodea <- with(data6, ifelse(wasInElodea=="I DID NOT operate a boat in any of the waterbodies listed", 0, ifelse(wasInElodea ==""|wasInElodea=="I cannot remember", NA, 1)))

mapResponse <- subset(data6, mapres==1)
mapNonRes <- subset(data6, mapres==0)

#Summary of respondents with mapping response
mapResSummary <- mapResponse%>%
  group_by(operateMarine,operateFresh)%>%
  summarise(median.min = round(median(Duration.s)/60, digits = 1),
            count = n(),
            elodeaCount = sum(Elodea, na.rm = TRUE))

#Summary of respondents without mapping response
non.mapResSummary <- mapNonRes%>%
  group_by(operateMarine,operateFresh)%>%
  summarise(median.min = round(median(Duration.s)/60, digits = 1),
            count = n(),
            elodeaCount = sum(Elodea, na.rm = TRUE))

#Rank by respondent frequency of elodea infestation recorded to be visited
locationSummary <- read.csv(textConnection(data6[["wasInElodea"]]), stringsAsFactors = FALSE)
names(locationSummary)[names(locationSummary)=="I.DID.NOT.operate.a.boat.in.any.of.the.waterbodies.listed"] <- "location" #renaming the column
locationSummary$location <- with(locationSummary, ifelse(location=="I DID NOT operate a boat in any of the waterbodies listed", "was not in elodea", ifelse(location=="I cannot remember","cannot remember",location)))

locationSummary<-locationSummary%>%
  group_by(location)%>%
  summarise(count=n())
library(knitr)
kable(locationSummary, caption ="Respondent count to known elodea locations")
# Other info about the sample and surveying
otherInfo <- effsample%>%
  summarise(onlineDiffRate = sum(onlineDifficulty)/nrow(effsample),
            noInternetRate = sum(noInternet/nrow(effsample)),
            returnedDollar = sum(returnIncentive),
            removeList = sum(requestsRemove))

# online difficulty 0.5%
# no Internet 0.3%
# 21 returned dollar incentive (answered)
# 2 requested to be removed from list (counted as non-respondent)

```


#Map of respondents and where they reside color coded by whethe repondents gave mapping answers/did not boat
Location Lat Long only available for 848 of the 965 respondents!!! 
```{r}
vars <- c("mapres","responseID","LocationLongitude.x","LocationLatitude.x","didNotOperate")
respondentLocations <- data6[vars]
#eliminting respondents where we could not get location information
respondentLocations <- subset(respondentLocations, !is.na(LocationLongitude.x))
respondentLocations$marker <- with(respondentLocations, ifelse(mapres==1, 1, ifelse(mapres==0&didNotOperate==0,0,2)))


getColor <- function(respondentLocations) {
  sapply(respondentLocations$marker, function(marker) {
  if(marker==1) {
    "green"
  } else if(marker==0) {
    "red"
  } else {
    "blue"
  } })
}

icons <- awesomeIcons(
  icon = 'ios-close',
  iconColor = 'black',
  library = 'ion',
  markerColor = getColor(respondentLocations)
)

leaflet(respondentLocations)%>%
    addTiles()%>%
    addProviderTiles(providers$Esri.WorldTopoMap)%>%
    addAwesomeMarkers(~LocationLongitude.x, ~LocationLatitude.x, icon=icons, label=~as.character(mapres))%>%
    addLegend("topright", colors = c("green","red", "blue"), labels = c("gave map response","no map response","did not boat"))

```

#Costs, trip frequency, number of routes, annual operating cost, BY USER GROUP
```{r}
#Creating descriptive stats tables
Hmisc::describe(features5$hCost )
Hmisc::describe(features5$YrCostPre)

tableData <- dplyr::select(features5, "hCost", "totalKmPre","totalKmPost", "km", "freqPre", "freqPost", "primaryPur")
    
stargazer::stargazer(tableData, digits=0,  out="tables/table1.txt")
stargazer::stargazer(tableData, digits=0, type = "html", out="tables/table1.html")
```

#Exploratory data analysis
Table showing main purpose of route by mean ranking
```{r}
StatsByPur <- features5 %>%
  dplyr::group_by(primaryPur)%>%
  summarise(percentRoutes = n()/nrow(features5),
            meanRank = mean(rank),
            count = n())%>%
  arrange(desc(count))
  
write.table(StatsByPur, file = "tables/table2.txt", sep = ",", quote = FALSE, row.names = F)

#getting density data, 9 commercial operators, 15 government, and 741 personal/private
c <- density(com$hCost)
g <- density(gov$hCost)
p <- density(pers$hCost)

#plotting kernel densities
plot(c)
plot(g)
plot(p)

# correlation between income and annual travel cost for personal operators

plot(pers$age, pers$Income)
```
















#High priority TASKS

2) can I visualize the frequencies on map? 
3) visualize out of state origin tracks

# Low priority tasks
3) Does a hex map make sense?
























# Integrating dynamic with static elements (Source: https://rstudio.github.io/leaflet/shiny.html)
```{r}
library(shiny)
library(leaflet)
library(RColorBrewer)

ui <- bootstrapPage(
  tags$style(type = "text/css", "html, body {width:100%;height:100%}"),
  leafletOutput("map", width = "100%", height = "100%"),
  absolutePanel(top = 10, right = 10,
    checkboxInput("select","Show tracks of",
      list(`boats operated outside Alaska before coming to Alaska` = list("NY", "NJ", "CT"),
           `boats Operated only in Alaska in 2018` = list(""),
           `all boats` = list("British Columbia","Washington","Kansas","Yukon Territories",""),
                selected = ""
           ),
      
        ),
    checkboxInput("legend", "Show legend", TRUE)
  )
)

```


```{r}
server <- function(input, output, session) {

  # Reactive expression for the data subsetted to what the user selected
  filteredData <- reactive({
    mapdata[mapdata$freshOutside =="Yes",]
  })

  # This reactive expression represents the palette function,
  # which changes as the user makes selections in UI.
  colorpal <- reactive({
    colorNumeric(input$colors, quakes$mag)
  })

  output$map <- renderLeaflet({
    # Use leaflet() here, and only include aspects of the map that
    # won't need to change dynamically (at least, not unless the
    # entire map is being torn down and recreated).
    leaflet(mapdata) %>% addTiles() %>%
      fitBounds(~min(long), ~min(lat), ~max(long), ~max(lat))
  })

  # Incremental changes to the map (in this case, replacing the
  # circles when a new color is chosen) should be performed in
  # an observer. Each independent set of things that can change
  # should be managed in its own observer.
  observe({
    pal <- colorpal()

    leafletProxy("map", data = filteredData()) %>%
      clearShapes() %>%
      addCircles(radius = ~10^mag/10, weight = 1, color = "#777777",
        fillColor = ~pal(mag), fillOpacity = 0.7, popup = ~paste(mag)
      )
  })

  # Use a separate observer to recreate the legend as needed.
  observe({
    proxy <- leafletProxy("map", data = quakes)

    # Remove any existing legend, and only if the legend is
    # enabled, create a new one.
    proxy %>% clearControls()
    if (input$legend) {
      pal <- colorpal()
      proxy %>% addLegend(position = "bottomright",
        pal = pal, values = ~mag
      )
    }
  })
}

shinyApp(ui, server)
 
``` 

