---
title: "boat movement analysis"
author: "Tobias Schwoerer"
date: "November 21, 2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Importing the geojson file containing the routes by respondent and then importing the survey data by respondent

```{r}
library(jsonlite)
library(geojson)
library(leaflet)
library(dplyr)
library(geojsonio)
#routes <- geojsonio::geojson_read("D:/Dropbox/CURRENT_PROJECTS/Boatmovements/Data/outputFinal.geojson", what = "sp")
#routes <- geojsonio::geojson_read("data/outputFinal.geojson", what = "sp")
#routes <- geojsonio::geojson_read("data/Boat_Survey_Nov_2018.json", what = "sp")
data <- geojson_read("D:/Dropbox/DATA/2018_boat_survey/map-data-joiner-output.geojson",what = "sp")
#geojsonlint::geojson_lint("data/Boat_Survey_Nov_2018.json.json")

#routes <- system.file("examples", "bison.zip", package = "geojsonio")
class(data)
answers <- read.csv("D:/Dropbox/DATA/2018_boat_survey/October22FINALdownload_clean_deIDed2.csv")
#answers <- read.csv("data/October22FINALdownload_clean_deIDed.csv")
```

#Writing answers into a Json file then combining the routes with the answers file by responseID
 "python3 map-data-joiner.py October22FINALdownload_clean_deIDed2.csv Boats_survey_Nov_2018_2.geojson output.geojson" in the case of the files you sent over to me. The expected parameters are survey responses, GeoJSON for joining, and whatever name you'd like the output GeoJSON to use.
```{r}
library(reticulate)
source_python("D:/Dropbox/DATA/2018_boat_survey/map-data-joiner.py", envir = parent.frame(), convert = TRUE) 
 
 
``` 
 
```{r}
# write survey answers as geojson
geojson_write(answers, "data/answers.geojson")

#validating this file as a true json file
geojsonlint::geojson_lint("data/answers.geojson")

answersJson <- geojson_read("data/answers.geojson", what = "sp")

# combining json files
files <- c("routes", "answersJson")
jsonl <- lapply(files, function(f) fromJSON(file = f))
jsonc <- toJSON(jsonl)
write(jsonc, file = "jsonc")
class(jsonc)

#creating leaflet map
leaflet(data) %>%
  addTiles() %>%
  addPolylines() 
```

#Analysis of other non-spatial survey answers
```{r}


```
