---
title: "boat movement analysis"
author: "Tobias Schwoerer"
date: "November 21, 2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Importing the data files, visualizing all respondent boat tracks 

```{r}
library(jsonlite)
library(geojson)
library(leaflet)
library(dplyr)
library(geojsonio)
library(tidyr)
library(Hmisc)
library(ggplot2)
library(RColorBrewer)

#cleaned up features (tracks the respondents entered into the mapping tool)
#map <- geojson_read("D:/Dropbox/DATA/2018_boat_survey/map.geojson",what = "sp")
map <- geojson_read("C:/Users/Toby/Dropbox/DATA/2018_boat_survey/map.geojson", what = "sp")

#features <- read.csv("D:/Dropbox/DATA/2018_boat_survey/features.csv", stringsAsFactors = FALSE)
features <- read.csv("C:/Users/Toby/Dropbox/DATA/2018_boat_survey/features.csv", stringsAsFactors = FALSE)

#renaming misspelled column
names(features)[names(features)=="LenghtKM"] <- "km"


freqCoding <- read.csv("reference_tables/freqCoding.csv", stringsAsFactors = FALSE)

#cleaned data associated with the survey answers of each respondent
#data <- read.csv("D:/Dropbox/DATA/2018_boat_survey/data.csv", stringsAsFactors = FALSE)
#data <- read.csv("C:/Users/Toby/Dropbox/DATA/2018_boat_survey/data.csv", stringsAsFactors = FALSE)

# the joined data and map output, joined these files using the python script
#mapdata <- geojson_read("D:/Dropbox/DATA/2018_boat_survey/output.geojson",what = "sp")
mapdata <- geojson_read("C:/Users/Toby/Dropbox/DATA/2018_boat_survey/output.geojson",what = "sp")

#creating leaflet map showing all routes
leaflet(mapdata) %>%
  addTiles() %>%
  addPolylines() %>%
  addProviderTiles(providers$Stamen.TerrainBackground)
```

#Calculating trip frequencies pre and post hypothetical elodea invasions of waterways used 
```{r}
#calculating pre elodea trip frequencies
features2 <- features%>%
  left_join(freqCoding,  by = c("trackFrequ"="Coding"))
features2$freqPre <- rowMeans(features2[c("Lower","Upper")], na.rm=TRUE)
#Renaming columns 
names(features2)[names(features2)=="Lower"] <- "LowerPre"
names(features2)[names(features2)=="Upper"] <- "UpperPre"

#Calculating post elodea trip frequencies
features3 <- features2 %>%
  left_join(freqCoding,  by = c("trackFre_1"="Coding"))
features3$freqPost <- rowMeans(features3[c("Lower","Upper")], na.rm=TRUE)
#Renaming columns 
names(features3)[names(features3)=="Lower"] <- "LowerPost"
names(features3)[names(features3)=="Upper"] <- "UpperPost"
```

#Outliers and missing data for features
#Calculating annual cost of boating per route assumes average travel speed of 10 miles/h equal to 16km/h
```{r}
#calculating mean cost from the high and low cost per hour respondents reported
#last argument excludes missing values
features3$hCost <- rowMeans(features3[c("hCostHigh","hCostLow")], na.rm=TRUE)

#dealing with outlier on cost
features3$hCost[features3$hCost==20000] <- NA
hist(features3$hCost, breaks=20)

#dealing with outliers on km, setting frequency of trips pre if Up to 4 setting it to 1 if km >300
features3$freqPre <- with(features3, ifelse(km>300 & trackFrequ=="Up to 4",1,freqPre))

#dealing with ranks that are 0
features3$rank <- with(features3, ifelse(rank==0, 1, rank)) 

#dealing with missing values by first calculating the median cost per h then inserting missing values with the median
medianCost <- median(features3$hCost, na.rm=TRUE)
features3$hCost <- with(features3, ifelse(is.na(hCost), medianCost, hCost))

#calculating the total km per route then annual cost per route
features3$totalKmPre <- with(features3, km * freqPre)
features3$totalKmPost <- with(features3, km * freqPost)
features3$YrCostPre <- with(features3, totalKmPre * hCost)
features3$YrCostPost <- with(features3, totalKmPost * hCost)
```

#cleaning up features further, missing data etc. 
```{r}
#getting rid of % sign in character string, then converting character to numeric, doing for percentCom and percentGov
features4 <- features3 
features4$percentCom <- gsub("[[:punct:]]", " ", features4$percentCom)
features4$percentCom <- as.numeric(as.character(features4$percentCom))
features4$percentCom <- with(features4, ifelse(is.na(percentCom),0,percentCom/100))
features4$percentGov <- gsub("[[:punct:]]", " ", features4$percentGov)
features4$percentGov <- as.numeric(as.character(features4$percentGov))
features4$percentGov <- with(features4, ifelse(is.na(percentGov),0,percentGov/100))
features4$percentPer <- with(features4, 1-percentGov-percentCom)

#correcting wrong respondent entry on percentGov
features4$percentGov <- with(features4, ifelse(percentPer<0,percentGov-percentCom,percentGov))
features4$percentPer <- with(features4, 1-percentGov-percentCom)

#renaming columns
names(features4)[names(features4)=="CleanDrainDry"] <- "Clean"

#turning character to numeric for income variable
incCoding <- read.csv("reference_tables/incCoding.csv", stringsAsFactors = FALSE)
features4 <- features4%>%
  dplyr::left_join(incCoding,  by = "persIncomeBtax")
features4$Income <- as.numeric(as.character(features4$Income))

#dealing with missing data on income
#dealing with missing values by first calculating the median income then inserting missing values with the median
medianIncome <- median(features4$Income, na.rm=TRUE)
features4$Income <- with(features4, ifelse(is.na(Income), medianIncome, Income))

#dealing with questionable data on age and deleting entries coded "Not applicable"
features4$age <- with(features4, ifelse(age==9 | age=="Not applicable",NA,age))
features4$age <- as.numeric(as.character(features4$age))

#coding respondents who did not respond to Clean Drain Dry question
features4$Clean <- with(features4, ifelse(Clean=="", "Did not report", ifelse(Clean=="half the time","50% of the itme",ifelse(Clean=="Every time", "100% of the time",ifelse(Clean=="Never", "0% of the time", Clean)))))

#dealine with miscoding
features4$Clean <- with(features4, ifelse(Clean=="50% of the itme", "50% of the time",Clean))

#turning character vectors into factors for Clean variable, for later when we need a nice legend for the plot
Features4$Clean <- factor(features4$Clean, levels=c("100% of the time", "75% of the time", "50% of the time", "25% of the time", "0% of the time", "Did not report"), labels=c("100% of the time", "75% of the time", "50% of the time", "25% of the time", "0% of the time", "Did not report"))

#cleaning up columns no longer needed
features5 <- subset(features4, select = -c(trackFrequ, trackFre_1, IPAddress, operateType, persIncomeBtax, LowerPost, LowerPre, UpperPost, UpperPre))

#Creating bins for age
features5$ageBin <- with(features5, ifelse(age>20&age<=30,"21-30",ifelse(age>30&age<= 40,"31-40",ifelse(age>40&age<=50,"41-50", ifelse(age>50&age<=60,"51-60", ifelse(age>60&age<=70,"61-70", ifelse(age>70&age<=80, "71-80", "81 and over")))))))
#turning ageBin into factor
features5$ageBin <- factor(features5$ageBin, levels=c("21-30", "31-40", "41-50", "51-60", "61-70", "71-80","81 and over"), labels=c("21-30", "31-40", "41-50", "51-60", "61-70", "71-80","81 and over"))

#creating subsets by operator type
#creating variable for personal and commercial operators
features5$type <- with(features5, ifelse(percentPer>0.5, "personal", ifelse(percentCom>0.5, "commercial", "government")))

# creating subsets for each type of operator
com <- subset(features5, type=="commercial")
pers <- subset(features5, type=="personal")
gov <- subset(features5, type=="government")
```

#Distance by personal operators by age





# hourly and yearly operating cost statistics (distributions)
```{r}
resSum <- features5%>%
  group_by(responseID, type)%>%
  summarise(YrTravelCost = sum(YrCostPre))

#Calculating descriptive stats for annual operating cost by operator type
TCByType <- resSum%>%
  group_by(type)%>%
  summarise(mean = mean(YrTravelCost),
            median = median(YrTravelCost),
            sd = sd(YrTravelCost),
            max = max(YrTravelCost),
            min = min(YrTravelCost),
            cv = sd/mean)

hCByType <- features5%>%
  group_by(type)%>%
  summarise(mean = mean(hCost),
            median = median(hCost),
            sd = sd(hCost),
            max = max(hCost),
            min = min(hCost),
            cv = sd/mean)
```


```{r}
#dealing with 50,000 total km of boat travel outlier
pers2 <- pers%>%
  subset(totalKmPre < 2000)
pers3 <- pers%>%
   Clean != "Did not report"

#turning character vectors into factors for Clean variable, for later when we need a nice legend for the plot
Features4$Clean <- factor(features4$Clean, levels=c("100% of the time", "75% of the time", "50% of the time", "25% of the time", "0% of the time", "Did not report"), labels=c("100% of the time", "75% of the time", "50% of the time", "25% of the time", "0% of the time", "Did not report"))


persAgeBin <- pers3 %>%
    filter(!is.na(ageBin)) %>%
    group_by(ageBin, Clean) %>%
    summarise(n = n())%>%
    spread(Clean, n)
persAgeBin[is.na(persAgeBin)] <- 0
#calculating row total
persAgeBin$total <- rowSums(persAgeBin[,2:6])

#renaming columns for easy mutate 
names(persAgeBin)[names(persAgeBin)=="25% of the time"] <- "n25Clean"
names(persAgeBin)[names(persAgeBin)=="50% of the time"] <- "n50Clean"
names(persAgeBin)[names(persAgeBin)=="75% of the time"] <- "n75Clean"
names(persAgeBin)[names(persAgeBin)=="0% of the time"] <- "n0Clean"
names(persAgeBin)[names(persAgeBin)=="100% of the time"] <- "n100Clean"

#calculating proportions
persAgeBin2 <- persAgeBin%>%
  mutate(p25 = round(n25Clean/total, digits=2),
         p50 = round(n50Clean/total, digits=2),
         p0 = round(n0Clean/total, digits=2),
         p75 = round(n75Clean/total, digits=2),
         p100 = round(n100Clean/total, digits=2))




persAgeBin <- pers3 %>%
    filter(!is.na(ageBin)) %>%
    group_by(ageBin, Clean) %>%
    summarise(n = n())

#distribution of annual km by age bin among the personal operators, just revise the y variable to km
ggplot(persAgeBin, aes(x = ageBin, y = n)) +
  geom_bar(stat = "identity", position = position_stack())+
  scale_color_manual(values = c("#0073C2FF", "#EFC000FF"))+
  scale_fill_manual(values = c("#0073C2FF", "#EFC000FF")) +theme_bw()

mean = mean(totalKmPre),
            median = median(totalKmPre),
            clean = n(totalKmPre),
            max = max(totalKmPre),
            min = min(totalKmPre),
            cv = sd/mean)


```

#Plotting the data
```{r}
#dealing with 50,000 total km of boat travel outlier
pers2 <- pers%>%
  subset(totalKmPre < 2000)
pers3 <- pers%>%
  subset(totalKmPre < 2000 &
         Clean != "Did not report")



#colors for plot
theme <- c('#a63603','#e6550d','#fd8d3c','#fdae6b','#fdd0a2','#feedde')
#plot
t <- ggplot(pers3, aes(age, totalKmPre, colour = Clean)) + geom_point() +
  geom_hline(aes(yintercept = 0)) + ylim(0,2000) 
t + labs(colour = "CleanDrainDry") +scale_color_manual(values=theme) +ylab("km / year")
```




#Clean Drain Dry behavior by age bin and operator type
In all of the above results the commercials are mixed in with the personal trips
# separating commercial from personal in data.csv
```{r}
#getting rid of % sign in character string, then converting character to numeric, doing for percentCom and percentGov
data2 <- data
data2$percentCom <- gsub("[[:punct:]]", " ", data2$percentCom)
data2$percentCom <- as.numeric(as.character(data2$percentCom))
data2$percentCom <- with(data2, ifelse(is.na(percentCom),0,percentCom/100))
data2$percentGov <- gsub("[[:punct:]]", " ", data2$percentGov)
data2$percentGov <- as.numeric(as.character(data2$percentGov))
data2$percentGov <- with(data2, ifelse(is.na(percentGov),0,percentGov/100))
data2$percentPer <- with(data2, 1-percentGov-percentCom)

#renaming columns
names(data2)[names(data2)=="CleanDrainDry"] <- "Clean"


#turning character to numeric for income variable
incCoding <- read.csv("reference_tables/incCoding.csv", stringsAsFactors = FALSE)
data3 <- data2%>%
  dplyr::left_join(incCoding,  by = "persIncomeBtax")
data3$Income <- as.numeric(as.character(data3$Income))

#dealing with missing data on income
#dealing with missing values by first calculating the median income then inserting missing values with the median
medianIncome <- median(data3$Income, na.rm=TRUE)
data3$Income <- with(data3, ifelse(is.na(Income), medianIncome, Income))

#creating a selection of columns from data to join to features
data3Select <- dplyr::select(data3, responseID, percentPer, percentCom, percentGov, Income, freshOutside, Clean)

```

#Plotting compliance with clean drain dry by age bin using the data.csv file with 965 observations not only the respondents who used the mapping tool
```{r}






persAgeBin <- pers3 %>%
    filter(!is.na(ageBin)) %>%
    group_by(ageBin, Clean) %>%
    summarise(n = n())  

#plotting proportions of clean drain dry proportions by age bin
library(ggplot2)
ggplot(persAgeBin, aes(x = ageBin, y = n, fill = Clean)) +
       geom_bar(stat = 'identity', position = 'fill', aes(fill = Clean))
```






#Exploratory data analysis
Table showing main purpose of route by mean ranking
```{r}
StatsByPur <- features5 %>%
  dplyr::group_by(primaryPur)%>%
  summarise(percentRoutes = n()/nrow(features5),
            meanRank = mean(rank),
            count = n())%>%
  arrange(desc(count))
  
write.table(StatsByPur, file = "tables/table2.txt", sep = ",", quote = FALSE, row.names = F)

#getting density data, 9 commercial operators, 15 government, and 741 personal/private
c <- density(com$hCost)
g <- density(gov$hCost)
p <- density(pers$hCost)

#plotting kernel densities
plot(c)
plot(g)
plot(p)

# correlation between income and annual travel cost for personal operators

plot(pers$age, pers$Income)

```




#Developing a color palette for printing 
```{r}
# Function for plotting colors side-by-side
pal <- function(col, border = "light gray", ...){
  n <- length(col)
  plot(0, 0, type="n", xlim = c(0, 1), ylim = c(0, 1),
       axes = FALSE, xlab = "", ylab = "", ...)
  rect(0:(n-1)/n, 0, 1:n/n, 1, col = col, border = border)
}

# FOCUS PALETTES
# Red as highlight
#redfocus = c("#CB181D", "#252525", "#525252", "#737373", "#969696", "#BDBDBD", "#D9D9D9", "#F0F0F0")
 redfocus = c("#CB181D", "#252525", "#525252", "#737373", "#969696", "#BDBDBD")
# Green as highlight
#greenfocus = c("#41AB5D", "#252525", "#525252", "#737373", "#969696", "#BDBDBD", "#D9D9D9", "#F0F0F0")
 greenfocus = c("#41AB5D", "#252525", "#525252", "#737373", "#969696", "#BDBDBD")
# Blue as highlight
bluefocus = c("#0033FF", "#252525", "#525252", "#737373", "#969696", "#BDBDBD", "#D9D9D9", "#F0F0F0")
 
pal(greenfocus)

# EQUAL WEIGHT
# Generated with rainbow(12, s = 0.6, v = 0.75)
rainbow12equal = c("#BF4D4D", "#BF864D", "#BFBF4D", "#86BF4D", "#4DBF4D", "#4DBF86", "#4DBFBF", "#4D86BF", "#4D4DBF", "#864DBF", "#BF4DBF", "#BF4D86")
rainbow10equal = c("#BF4D4D", "#BF914D", "#A8BF4D", "#63BF4D", "#4DBF7A", "#4DBFBF", "#4D7ABF", "#634DBF", "#A84DBF", "#BF4D91")
rainbow8equal = c("#BF4D4D", "#BFA34D", "#86BF4D", "#4DBF69", "#4DBFBF", "#4D69BF", "#864DBF", "#BF4DA3")
rainbow6equal = c("#BF4D4D", "#BFBF4D", "#4DBF4D", "#4DBFBF", "#4D4DBF", "#BF4DBF")
 
# Generated with package "gplots" function rich.colors(12)
rich12equal = c("#000040", "#000093", "#0020E9", "#0076FF", "#00B8C2", "#04E466", "#49FB25", "#E7FD09", "#FEEA02", "#FFC200", "#FF8500", "#FF3300")
rich10equal = c("#000041", "#0000A9", "#0049FF", "#00A4DE", "#03E070", "#5DFC21", "#F6F905", "#FFD701", "#FF9500", "#FF3300")
rich8equal = c("#000041", "#0000CB", "#0081FF", "#02DA81", "#80FE1A", "#FDEE02", "#FFAB00", "#FF3300")
rich6equal = c("#000043", "#0033FF", "#01CCA4", "#BAFF12", "#FFCC00", "#FF3300")
 
# Generated with package "fields" function tim.colors(12), which is said to emulate the default matlab colorset
tim12equal = c("#00008F", "#0000EA", "#0047FF", "#00A2FF", "#00FEFF", "#5AFFA5", "#B5FF4A", "#FFED00", "#FF9200", "#FF3700", "#DB0000", "#800000")
tim10equal = c("#00008F", "#0000FF", "#0070FF", "#00DFFF", "#50FFAF", "#BFFF40", "#FFCF00", "#FF6000", "#EF0000", "#800000")
tim8equal = c("#00008F", "#0020FF", "#00AFFF", "#40FFBF", "#CFFF30", "#FF9F00", "#FF1000", "#800000")
tim6equal = c("#00008F", "#005AFF", "#23FFDC", "#ECFF13", "#FF4A00", "#800000")
 
# Generated with sort(brewer.pal(8,"Dark2")) #Dark2, Set2
dark8equal = c("#1B9E77", "#666666", "#66A61E", "#7570B3", "#A6761D", "#D95F02", "#E6AB02", "#E7298A")
dark6equal = c("#1B9E77", "#66A61E", "#7570B3", "#D95F02", "#E6AB02", "#E7298A")
set8equal = c("#66C2A5", "#8DA0CB", "#A6D854", "#B3B3B3", "#E5C494", "#E78AC3", "#FC8D62", "#FFD92F")
set6equal = c("#66C2A5", "#8DA0CB", "#A6D854", "#E78AC3", "#FC8D62", "#FFD92F")
 
pal(rich8equal)

```
















#Some exploratory analysis
```{r}
#Creating descriptive stats tables
Hmisc::describe(features3$hCost )
Hmisc::describe(features3$YrCostPre)

tableData <- dplyr::select(features4, "hCost", "totalKm", "km", "freqPre", "freqPost", "primaryPur")
    
stargazer::stargazer(tableData, digits=0,  out="tables/table1.txt")
stargazer::stargazer(tableData, digits=0, type = "html", out="tables/table1.html")
```










#High priority TASKS

2) can I visualize the frequencies on map? 
3) visualize out of state origin tracks

# Low priority tasks
3) Does a hex map make sense?
























# Integrating dynamic with static elements (Source: https://rstudio.github.io/leaflet/shiny.html)
```{r}
library(shiny)
library(leaflet)
library(RColorBrewer)

ui <- bootstrapPage(
  tags$style(type = "text/css", "html, body {width:100%;height:100%}"),
  leafletOutput("map", width = "100%", height = "100%"),
  absolutePanel(top = 10, right = 10,
    checkboxInput("select","Show tracks of",
      list(`boats operated outside Alaska before coming to Alaska` = list("NY", "NJ", "CT"),
           `boats Operated only in Alaska in 2018` = list(""),
           `all boats` = list("British Columbia","Washington","Kansas","Yukon Territories",""),
                selected = ""
           ),
      
        ),
    checkboxInput("legend", "Show legend", TRUE)
  )
)

```


```{r}
server <- function(input, output, session) {

  # Reactive expression for the data subsetted to what the user selected
  filteredData <- reactive({
    mapdata[mapdata$freshOutside =="Yes",]
  })

  # This reactive expression represents the palette function,
  # which changes as the user makes selections in UI.
  colorpal <- reactive({
    colorNumeric(input$colors, quakes$mag)
  })

  output$map <- renderLeaflet({
    # Use leaflet() here, and only include aspects of the map that
    # won't need to change dynamically (at least, not unless the
    # entire map is being torn down and recreated).
    leaflet(mapdata) %>% addTiles() %>%
      fitBounds(~min(long), ~min(lat), ~max(long), ~max(lat))
  })

  # Incremental changes to the map (in this case, replacing the
  # circles when a new color is chosen) should be performed in
  # an observer. Each independent set of things that can change
  # should be managed in its own observer.
  observe({
    pal <- colorpal()

    leafletProxy("map", data = filteredData()) %>%
      clearShapes() %>%
      addCircles(radius = ~10^mag/10, weight = 1, color = "#777777",
        fillColor = ~pal(mag), fillOpacity = 0.7, popup = ~paste(mag)
      )
  })

  # Use a separate observer to recreate the legend as needed.
  observe({
    proxy <- leafletProxy("map", data = quakes)

    # Remove any existing legend, and only if the legend is
    # enabled, create a new one.
    proxy %>% clearControls()
    if (input$legend) {
      pal <- colorpal()
      proxy %>% addLegend(position = "bottomright",
        pal = pal, values = ~mag
      )
    }
  })
}

shinyApp(ui, server)
 
``` 
 
```{r}
# write survey answers as geojson
geojson_write(answers, "data/answers.geojson")

#validating this file as a true json file
geojsonlint::geojson_lint("data/answers.geojson")

answersJson <- geojson_read("data/answers.geojson", what = "sp")

# combining json files
files <- c("routes", "answersJson")
jsonl <- lapply(files, function(f) fromJSON(file = f))
jsonc <- toJSON(jsonl)
write(jsonc, file = "jsonc")
class(jsonc)

#creating leaflet map
leaflet(data) %>%
  addTiles() %>%
  addPolylines() 
```

