---
title: "Study of freshwater AIS pathway: Alaska boat ownership"
author: "Tobias Schwoerer"
date: "November 29, 2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

##1. Sample frame 
The sample frame was created with a data dump from the Alaska Department of Motorvehicles (ADMV) registration database as of December 2017. The database contains any current and historic registered boats including first owners' address , multi-owners' names, and information about boat characteristics including boat length, engine type and power, etc. The data dump was received through Tammy Davis at the Alaska Department of Fish and Game, the study's sponsor. 

The data was cleaned of any space occurring after text strings, columns were renamed for shorter code, and ZIP codes were trimmed. Registered house boats and commercial fishing boats were not included in the sample frame. A subset of this database was then further used that showed only the currently registered boats with registration expiring in 2018 or later. In order to test whether a stratified random sample was preferred over a simple random sample, we included information on geographic regions, approximately consistent with Alaska's HUC6 watersheds.  
```{r include=FALSE}

library(dplyr)
library(tidyr)
library(stringr)
library(sampling)
library(kableExtra)

#Importing dataset, renaming columns
hisReg <- read.csv("D:/Dropbox/DATA/2018_boat_survey/registration/DMV_registered_boats_data_2017.csv", stringsAsFactors=FALSE)
#hisReg <- read.csv("C:/Users/Toby/Dropbox/DATA/2018_boat_survey/registration/DMV_registered_boats_data_2017.csv", stringsAsFactors=FALSE)

register <- hisReg
#renaming columns
names(register)[names(register)=="ADDR.RES.CITY"] <- "Community"
names(register)[names(register)=="OWNRSHIP.TYPE.1"] <- "OWNER1"
names(register)[names(register)=="OWNRSHIP.TYPE.2"] <- "OWNER2"
names(register)[names(register)=="OWNRSHIP.TYPE.3"] <- "OWNER3"
names(register)[names(register)=="OWNRSHIP.TYPE.4"] <- "OWNER4"

#remove all white space in the file, including in the column headers
names(register)<-names(register)%>% 
  stringr::str_replace_all("\\s","_")
register$Community <- str_trim(register$Community)
register$OWNER.NAME.FIRST.1 <- str_trim(register$OWNER.NAME.FIRST.1)
register$OWNER.NAME.LAST.1 <- str_trim(register$OWNER.NAME.LAST.1)
register$OWNER.NAME.LAST.2 <- str_trim(register$OWNER.NAME.LAST.2)
register$OWNER.NAME.LAST.3 <- str_trim(register$OWNER.NAME.LAST.3)
register$OWNER.NAME.LAST.4 <- str_trim(register$OWNER.NAME.LAST.4)
register$OWNER.NAME.MIDDLE.1 <- str_trim(register$OWNER.NAME.MIDDLE.1)
register$CATEGORY.SUB <- str_trim(register$CATEGORY.SUB)
register$USE.TYPE <- str_trim(register$USE.TYPE)

#ADDR.MAIL.ZIP column needs to be trimmed to five digits
register$ADDR.MAIL.ZIP <- strtrim(register$ADDR.MAIL.ZIP,5)
register$OWNER.NAME.COMPANY.1 <- str_trim(register$OWNER.NAME.COMPANY.1)

# Turn middle names into just one capital letter
register$OWNER.NAME.MIDDLE.1 <- substring(register$OWNER.NAME.MIDDLE.1, 1, 1)
register$NAME1 <- paste(register$OWNER.NAME.FIRST.1, register$OWNER.NAME.MIDDLE.1, register$OWNER.NAME.LAST.1, register$OWNER.NAME.COMPANY.1, sep=" ")
register$NAME2 <- paste(register$OWNER.NAME.FIRST.2, register$OWNER.NAME.MIDDLE.2, register$OWNER.NAME.LAST.2, register$OWNER.NAME.COMPANY.2, sep=" ")
register$NAME3 <- paste(register$OWNER.NAME.FIRST.3, register$OWNER.NAME.MIDDLE.3, register$OWNER.NAME.LAST.3, register$OWNER.NAME.COMPANY.3, sep=" ")
register$NAME4 <- paste(register$OWNER.NAME.FIRST.4, register$OWNER.NAME.MIDDLE.4, register$OWNER.NAME.LAST.4, register$OWNER.NAME.COMPANY.4, sep=" ")
register$multi_own <- with(register, ifelse((OWNER2=="O" & OWNER3==" "),2,ifelse((OWNER3=="O" & OWNER4==" "),3,ifelse((OWNER4=="O"|OWNER4=="S"),4,1))))

#joining region aggregation by using reference table
regions <-  read.csv("reference_tables/index_table_051118.csv", stringsAsFactors=FALSE)
names(regions)[names(regions)=="SASAPRegions"] <- "Regions"
regions <- subset(regions,select= c("Community", "Regions","Urban"))
regions <- mutate_at(regions,vars(Community), funs(toupper))
register2 <- register %>%
  left_join(regions, by="Community")
register2[is.na(register2)]<- "Outside Alaska"

#Excluding houseboats CATEGORY.SUB == D, commercial fishing boats USE.TYPE == C
register3 <- subset(register2, USE.TYPE!="C") # taking out fishing boats
register4 <- subset(register3, CATEGORY.SUB!="D") #taking out houseboats

#Subset for current registrations only
register2018 <- subset(register4, REG.YR.EXPIRE >=2018)
```

There are 54809 registered boats in the state of Alaska with registrations expiring wither in 2018 or later. Since a boat owner can have multiple boats, we created a set of unique last names and street addresses associated with owners who are listed first in titles and consequently the database. There are 45401 boat owners first listed on titles. 

```{r echo=FALSE}
#Unique first owners name (incl. company) and unique street address field
boat_owners <- register2018 %>%
  distinct(NAME1, ADDR.MAIL.STREET, .keep_all=TRUE)
```

Accounting for first, second, third and forth owners, we again created a list of unique names totalling  56103 owners, therefore 10702 owners are listed either as second, third, or forth owners. 
```{r}
allOwners_long <- stack(register2018, 
                     select=c(NAME1, 
                              NAME2, 
                              NAME3,
                              NAME4))%>%
  distinct(values)
totalOwnership <- unique(allOwners_long[, "values"])
```  


###1.1 Ownership and boat types 
Boat types recorded in the ADMV registration database include open skiffs, cabin cruisers, personal water craft (e.g. jet skis), sailboats, and a catch all category. Below we show frequency counts for boat types registered by region where the first boat owner listed in the database (in case of multi-ownership) resided in. While each Alaska boat owner was grouped into one of Alaska's large watershed regions,  boat owners residing outside Alaska were grouped into an Outside category. 
```{r echo=FALSE}
boatTypesByRegion <- register2018 %>%
  group_by(Regions)%>%
  summarise(Skiffs = sum(CATEGORY.SUB=="A"),
            Cabin.Cruisers = sum(CATEGORY.SUB =="B"),
            Personal.Water.Craft = sum(CATEGORY.SUB =="C"),
            Sailboats = sum(CATEGORY.SUB =="E"),
            other = sum(CATEGORY.SUB =="F")
            )
boatTypesByRegion$total <- with(boatTypesByRegion,rowSums(boatTypesByRegion[c("Skiffs","Cabin.Cruisers","Personal.Water.Craft","Sailboats","other")]))
boatPerc <- boatTypesByRegion %>%
   mutate(Skiffs_pct = round(Skiffs/total*100,digits=1),
            Cabin.Cruisers_pct = round(Cabin.Cruisers/total*100,digits=1),
            PWC_pct = round(Personal.Water.Craft/total*100,digits=1),
            Sail.Boats_pct = round(Sailboats/total*100,digits=1),
            Other_pct = round(other/total*100, digits=1))
boatPerc <- subset(boatPerc, select = -c(Skiffs, Cabin.Cruisers, Personal.Water.Craft, Sailboats, other, total)) #dropping unnecessary columns
knitr::kable(boatPerc, caption = "Percent of boats owned by type and region of owner residence, 2018")%>%
  kable_styling()

# registered boats per owner on average by type
ownersByRegion <- boat_owners%>%
  group_by(Regions)%>%
  summarise(Owners = n())

Region_centroids <- read.csv("reference_tables/Region_centroids.csv", stringsAsFactors = FALSE)
Region_centroids$Region <- with(Region_centroids, ifelse(Region=="Aleutian Islands", "Alaska Peninsula and Aleutian Islands", Region))
#cleaning up columns no longer needed
Region_centroids <- subset(Region_centroids, select = -c(FID, region_id, area_km))

#Adding row for Outside Alaska lat 47.6062, -122.3321
add<-data.frame("Outside Alaska",-122.3321,47.6062 )
names(add)<-c("Region","long", "lat")

Region_centroids <- rbind(Region_centroids, add)

Ownership <- ownersByRegion%>%
  left_join(boatTypesByRegion, by="Regions")%>%
  left_join(Region_centroids, by=c("Regions"="Region"))%>%
  dplyr::mutate(SkiffsAvg = round(Skiffs/Owners,digits=2),
              CabinCruisersAvg = round(Cabin.Cruisers/Owners, digits=2),
              PersonalWaterCraftAvg = round(Personal.Water.Craft/Owners,digits=2),
              SailboatsAvg = round(Sailboats/Owners, digits=2),
              otherAvg = round(other/Owners, digits=2) )
Ownership$total <- with(Ownership,rowSums(Ownership[c("Skiffs","Cabin.Cruisers","Personal.Water.Craft","Sailboats","other")]))
Ownership$boatsPerOwner <- with(Ownership, round(total/Owners, digits=1))

#Table: Owner count, ownership by boat type and average number of boats owned in 2018
Ownership_short <- Ownership[,-c(9:16)]
Ownership_short%>%
janitor::adorn_totals("row")

knitr::kable(Ownership_short, caption = "Owner and boat count by boat type and region of owner residence, 2018")%>%
  kable_styling()
```

###1.2. Percentage of boat types by region of owner residence, 2018
```{r echo=FALSE}
library(leaflet)
library(leaflet.minicharts)
basemap <- leaflet(width = "100%", height = "650px") %>%
  addTiles()%>%
  addProviderTiles(providers$Esri.WorldTopoMap)

colors <- c("#d7191c","#fdae61","#ffffbf","#abd9e9","#2c7bb6")

basemap %>%
  addMinicharts(
    Ownership$long, Ownership$lat,
    type = "pie",
    chartdata = Ownership[, c("Skiffs","Cabin.Cruisers","Personal.Water.Craft","Sailboats","other")], 
    colorPalette = colors, 
    width = 50 , transitionTime = 0
  )
# width = 60 * sqrt(Ownership$total) / sqrt(max(Ownership$total)), transitionTime = 0
# historical trends in boat type frequency, see if there is an increase in personal water craft, can't look at this since database is showing expired and not renewed assets for a certain year and not what in that year was actually registered. In cany case below is the code. 
```


##2. Sample
Based on the above groupings and through external analysis (in Excel), we determined that the simple random sample without replacement would provide for the smallest margin of error compared to a stratified random sample. We drew a pretest sample of 50 boat owners which was later extracted from the sample frame when drawing the final sample of 5095 boat owners. Sample size was determined in external analysis in Excel.   

###2.1 Pretest
For the pretest, we selected a simple random sample withough replacement of sample size 50. 
```{r echo=FALSE}
##Creating simple random sample for pre-test
frame <- boat_owners
sam.srswor <- srswor(n=50,N=45401)
sample.srswor <- frame[which(x=(sam.srswor ==1)),]

##Creating mailout database for pretest
pretest <- select(sample.srswor, OWNER.NAME.LAST.1, OWNER.NAME.FIRST.1, OWNER.NAME.MIDDLE.1, OWNER.NAME.SFX.1, OWNER.NAME.COMPANY.1, ADDR.MAIL.STREET, ADDR.MAIL.EXTRA.LINE, ADDR.MAIL.CITY, ADDR.MAIL.STATE, ADDR.MAIL.COUNTRY, ADDR.MAIL.ZIP , PROPERTY.ISN, CATEGORY.SUB, USE.TYPE)
#write.csv(pretest, "G:/My Drive/CURRENT_PROJECTS/4_Boatmovements/pre_test/pretest_contacts.csv")
```

###2.2 Final contact lists for business owners and individuals
For the final study, we selected a simple random sample without replacement of size 5095. We then distinguished between businesses and private individuals and created two mail out lists. 
```{r echo=FALSE}
##Creating remaining frame, eliminating pre-test contacts
eliminate <- pretest$PROPERTY.ISN
frame2 <- subset(boat_owners, !(boat_owners$PROPERTY.ISN %in% eliminate))

##Creating simple random sample for first letter mail out
sam.srswor2 <- srswor(n=5095,N=45401-50)
sample.srswor2 <- frame2[which(x=(sam.srswor2 ==1)),]
mailout <- select(sample.srswor2, OWNER.NAME.LAST.1, OWNER.NAME.FIRST.1, OWNER.NAME.MIDDLE.1, OWNER.NAME.SFX.1, OWNER.NAME.COMPANY.1, ADDR.MAIL.STREET, ADDR.MAIL.EXTRA.LINE, ADDR.MAIL.CITY, ADDR.MAIL.STATE, ADDR.MAIL.COUNTRY, ADDR.MAIL.ZIP , PROPERTY.ISN, CATEGORY.SUB, USE.TYPE)

#Separating sample into a business and private sample
business <- subset(mailout, OWNER.NAME.COMPANY.1!="")
private <- subset(mailout, OWNER.NAME.LAST.1!="")
  
#write.csv(business, "G:/My Drive/CURRENT_PROJECTS/4_Boatmovements/sampling/business_mailout.csv")
#write.csv(private, "G:/My Drive/CURRENT_PROJECTS/4_Boatmovements/sampling/private_mailout.csv")

## Creating table of PROPERTY.ISN (serving as unique ID), and ZIP code, for research subject payment documentation
#table_lexi <- mailout[,c("PROPERTY.ISN","ADDR.MAIL.CITY","ADDR.MAIL.ZIP")]
#write.csv(table_lexi,"G:/My Drive/CURRENT_PROJECTS/4_Boatmovements/sampling/table_lexi.csv")
```

###2.3 Analysis of multi-ownership
```{r echo=FALSE}

```